---
title: "Exploratory plot 2"
author: "Minh Tri Ngo"
format: pdf
---

#Packages 
```{r}
library(tidyverse)
library(janitor)
library(ggrepel)
library(cowplot)
library(here)
library(ggtext)
library(flextable)
```

- to do: 
- add caption to fig 4
- change labels and data to remove sheep but keep fire
- add figure captions and annotations
- do the gen ai declaration google form 
- do final check

# Data/Cleaning
```{r}
weather <- read_csv(here("data", "raw", "NOAA-weather-data.csv"))
# creating new object from weather
weather_clean <- weather |> 
  # cleaning column names
  clean_names() |> 
  # making sure date is in POSIXct format
  mutate(date = mdy(date)) |> 
  # selecting columns of interest
  select(date, prcp, tmax, tmin) |> 
  # extracting month and year from date column
  mutate(month = month(date),
         year = year(date)) |> 
  # assigning water year
  mutate(wy = case_when(
    # when month is October or later, assign value of year + 1
    month >= 10 ~ year + 1,
    # when month is September or earlier, assign value of existing year
    .default = year))
```

# Calculating Total Mean Annual Precipitation
```{r}
 tmap <- weather_clean |>
  group_by(year) |>
  summarize(total_precip = sum(prcp, na.rm = TRUE)) |> 
  ungroup()
```

## Plotting

## Total Annual Precip vs Grand Mean % Cover Native
```{r}
TAPnat <- ggplot(
  
)
```

## Total Annual Precip vs Grand Mean % Non-Native Plants scatterplots (facet wrap?/patchwork)
```{r}
TAPnnat <- ggplot(
  aes = 
)
```

## Add a trendline to the plots
- y axis: native and non-native 