---
title: "Grassland Vegetation Dynamics Under Intentional Disturbance at North Campus Open Space"
author: "Lukas Lescano, Garrett Mellinger, Minh Tri Ngo"
format: pdf
bibliography: references.bib
csl: ecology.csl
---

```{r}
#| label: packages
#| message: false
#| warning: false
#| echo: false

# data wrangling and general use
library(tidyverse)
library(janitor)
library(here)

# plot enhancements
library(ggrepel)
library(cowplot)
library(ggtext)

# table formatting
library(flextable)

# saving tables
library(webshot2)
```

```{r}
#| label: data
#| message: false
#| echo: false

# read in veg data
veg <- read_csv(here("data", "raw", "veg.csv"))

# read in metadata
metadata <- read_csv(here("data", "raw", "vp_veg_metadata.csv"))
```

```{r}
#| label: adding target species list
#| message: false
#| echo: false

# all grass species (Poaceae) at NCOS mesa, both native and non-native
poaceae_species <- c(
  "Bromus_hordeaceus",
  "Festuca_perennis",
  "Stipa_pulchra",
  "Polypogon_monspeliensis",
  "Brachypodium_distachyon",
  "Festuca_myuros",
  "Avena_barbata",
  "Hordeum_marinum",
  "Avena_fatua",
  "Parapholis_incurva",
  "Bromus_diandrus",
  "Distichlis_spicata",
  "Bromus_catharticus",
  "Hordeum_murinum",
  "Hordeum_brachyantherum_brachy.",
  "Polypogon_viridis",
  "Cortaderia_selloana",
  "Festuca_bromoides",
  "Cynodon_dactylon",
  "Poa_annua",
  "Festuca_microstachys",
  "Crypsis_schoenoides",
  "Polypogon_interruptus",
  "Hordeum_brachyantherum",
  "Bromus_madritensis",
  "Elymus_triticoides"
)

# native forb species to exclude from grass analysis
native_forbs <- c(
  "Plantago_erecta",
  "Sisyrinchium_bellum",
  "Sidalcea_malviflora",
  "Eryngium_vaseyi",
  "Acmispon_maritimus",
  "Acmispon_wrangelianus",
  "Madia_sativa",
  "Psilocarphus_tennulus",
  "Psilocarphus_brevissimus",
  "Spergularia_macrotheca",
  "Centromadia_parryi_australis",
  "Lepidium_nitidum",
  "Stebbinsoseris_heterocarpa",
  "Laennecia_coulteri"
)

# non-vegetation cover and non-target species to exclude from analysis
exclude_other <- c(
  "Bare Ground",
  "Natural Thatch",
  "unlisted",
  "moss",
  "boulder_rock",
  "biocrust",
  "Baccharis_pilularis",
  "Juncus_bufonius",
  "Juncus_phaeocephalus",
  "Eleocharis_macrostachya",
  "Eleocharis_acicularis",
  "Eleocharis_palustris"
)
```


```{r}
#| label: creating species lookup
#| message: false
#| echo: false

# creating data frame from veg
species_lookup <- veg |>
  # keep only GL transects at ncos
  filter(str_detect(transect_name, "GL"), site == "ncos") |>
  # remove unlisted since it maps to multiple categories
  filter(psoc != "unlisted") |>
  # return one unique row per species-category combination
  distinct(psoc, cover_category) |>
  # sort alphabetically by species for readability
  arrange(psoc)
```


```{r}
#| label: creating explicit0 dataframe
#| message: false
#| echo: false

# create a new object from veg
gl_explicit0 <- veg |> 
  # filter all occurrences where transect_name contains VP
  filter(str_detect(transect_name, "GL"),
         site == "ncos") |> 
  # not including cover_category because we will join it back after complete()
  select(-cover_category) |> 
  # complete all possible combinations of year, pool, psoc and distance
  complete(year, transect_name, psoc, transect_distance,
           # fill values of percent cover with 0
           fill = list(percent_cover = 0)) |>
  # group by year, pool, and psoc
  group_by(year, transect_name, psoc) |>
  # calculate total percent cover across all quads
  summarize(sum_pc = sum(percent_cover, na.rm = TRUE)) |>
  # ungroup the data frame
  ungroup() |>
  # create a new column called year_pool from year and transect name
  unite("year_pool", year, transect_name, remove = TRUE) |> 
  # join with metadata data frame
  left_join(metadata, by = "year_pool") |> 
  # calculate mean percent cover from total percent cover/number of quadrats
  mutate(mean_pc = sum_pc/num_quad) |> 
  # re-attach cover_category from lookup using psoc as key
  left_join(species_lookup, by = "psoc") |> 
  # excluding "GL-11" because it was unburned
  filter(transect_name != "GL-11")
```

```{r}
#| label: calculating change in mean percent cover
#| message: false
#| echo: false
#| warning: false

## Change in percentage points of focal species percent cover
## 2023 = pre-burn
## 2024 = post-burn

# define focal species with display labels and native/invasive status
focal_species <- tibble(
  psoc = c(
    "Stipa_pulchra",
    "Medicago_polymorpha",
    "Melilotus_indicus",
    "Festuca_perennis"
  ),
  species_label = c(
    "S. pulchra*",
    "M. polymorpha",
    "M. indicus*",
    "F. perennis"
  ),
  status = c(
    "Native",
    "Invasive",
    "Invasive",
    "Invasive"
  )
)

# filter gl_explicit0 to focal species and pre/post burn years only
focal_cover <- gl_explicit0 |>
  filter(
    year %in% c(2023, 2024),
    psoc %in% focal_species$psoc
  ) |>
  # keep only columns needed for change calculation
  select(year, transect_name, psoc, mean_pc) |>
  # attach species labels and status from focal_species
  left_join(focal_species, by = "psoc")

# calculate per-transect percentage point change from 2023 to 2024
focal_change_transect_pp <- focal_cover |>
  # pivot to wide format so 2023 and 2024 are separate columns
  pivot_wider(
    names_from = year,
    values_from = mean_pc,
    names_prefix = "year_"
  ) |>
  # calculate difference in mean percent cover between years
  mutate(
    percent_point_change = year_2024 - year_2023
  ) |>
  # remove transects missing data for either year
  filter(!is.na(percent_point_change))

# aggregate to mean percentage point change per species across transects
focal_change_summary_pp <- focal_change_transect_pp |>
  # group by spp, spp label, and status
  group_by(psoc, species_label, status) |>
  # calculate mean percent point change
  summarize(
    mean_percent_point_change = mean(percent_point_change, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) |>
  # set factor levels to control species order in plots
  mutate(
    species_label = factor(
      species_label,
      levels = c(
        "S. pulchra*",
        "M. polymorpha",
        "M. indicus*",
        "F. perennis"
      )
    ),
    # set factor levels to control fill order in legend
    status = factor(status, levels = c("Native", "Invasive"))
  )
```

```{r}
#| label: plot 1 creating grasses dataframe
#| message: false
#| echo: false

# creating grasses data frame from gl_explicit0
grasses <- gl_explicit0 |> 
  # filtering to include poaceae grasses species
  filter(psoc %in% poaceae_species,
         # filtering to include only native and non-native cover categories
         cover_category %in% c("NATIVE COVER", "NON-NATIVE COVER"))
```

```{r}
#| label: plot 1 creating grasses_summary for time series plot
#| message: false
#| echo: false

grasses_summary <- grasses |> 
  # grouping by year and cover category
  group_by(year, cover_category) |> 
  # calculating average per transect mean_pc across all transects and species within each group
  summarise(grand_mean_pc = mean(mean_pc, na.rm = TRUE),
            se = sd(mean_pc, na.rm = TRUE) / sqrt(n())) |> 
  # remove grouping
  ungroup()
```

```{r}
#| label: plot 1 creating grasses transect data frame
#| message: false
#| echo: false

# creating new object from grasses
grasses_transect <- grasses |>
  # grouping by year, transect, and conver category
  group_by(year, transect_name, cover_category) |>
  # calculating mean pc
  summarize(mean_pc = mean(mean_pc, na.rm = TRUE)) |>
  # removing groups
  ungroup()
```

# Introduction

This study assesses the progress of perennial grassland restoration on the North Campus Open Space (NCOS) Mesa using annual vegetation monitoring data from 2021 to 2025, with a focus on the immediate impact of intentional disturbance as a means of bolstering key native grass populations and suppressing exotic species abundance. NCOS is an active wetland restoration site with restored upland habitats such as native grassland, coastal sage scrub, and vernal pools (Rickard, 2024). Restoration success on the Mesa depends on both increasing native grass cover and understanding how non-native vegetation responds to restoration practices. Thus, we focus on native grass species as restoration targets and non-native grass and forb species as grassland invaders.

Prescribed fire is often used in California grasslands to manage competition because it can reduce exotic annual grasses, lower thatch cover, and create conditions that may favor native perennial species (Rickard, 2024). The fall 2023 cultural burn on the Mesa serves as an imperfect opportunity to evaluate whether fire can support restoration goals of shifting plant communities from exotic annual dominance toward native perennial grass recovery in this system and beyond. However, previous California grassland studies show that fire effects are not uniform across plant communities. Outcomes depend on which plant groups are present and when and how often burning occurs. For example, prescribed burn effects on California coastal prairie seed banks varied among native and non-native grasses and non-native forbs (Galloway et al., 2026), and burn timing and frequency influenced whether vegetation composition shifted in the intended direction (Hankins, 2013; Carlsen et al., 2017). Because native perennial grass recovery can take multiple years, we evaluate only short-term vegetation response one year after the 2023 cultural burn (Menke, 1992; Keeley et al., 2023). Specifically, we ask:

1. How does the percent cover of native grasses vary over time compared to the percent cover of non-native vegetation on the NCOS Mesa Perennial Grasslands?
    a. How does intentional disturbance influence this relationship?
    b. How do key individual species, including _Stipa pulchra_, _Medicago polymorpha_, _Melilotus indicus_, and _Festuca perennis_, vary in percent cover before and after intentional disturbance?

# Methods
  
Our analysis used the perennial grassland (Mesa) monitoring data from 2021 to 2025, with particular focus on 2023 and 2024, the years immediately before and after the burn. Grassland surveys were conducted annually in July. Focal species were selected based on overlap with the target species of previous sheep-grazing and burn-disturbance studies at NCOS. Additionally, our percent-change figure (Figure 2) was adapted from the percent biomass change approach used in the sheep-grazing study, substituting percent change in cover for biomass (Nestor, 2025).

Within the 16.8-acre Mesa grassland, short-low-growing vegetation is monitored along permanent 30-meter transects using one-square-meter quadrats (Rickard, 2024). Quadrats were placed every 3 meters, alternating between the left and right of the transect line, with the first quadrat centered to the left of the transect start. This produced 11 quadrats per transect and 88 quadrats across the 8 transects. Each quadrat was divided into ten 10-centimeter squares, and percent cover was estimated for every species present. For our study, we excluded GL-11 because it wasn't burned, which left us with 7 transects and 77 quadrats.

To address our third question, whether individual focal species changed in cover before and after the burn, we used paired t-tests comparing mean percent cover in 2023 (pre-burn) and 2024 (post-burn) for each of the four focal species: *Stipa pulchra*, *Medicago polymorpha*, *Melilotus indicus*, and *Festuca perennis*. Observations were paired by transect (n = 7), so each test evaluated whether the per-transect change in cover between the two years differed from zero. We assessed significance at α = 0.05 and did not apply a multiple-comparison correction, given the small number of pre-specified focal species. All tests were conducted in R using the t.test function.

To explore precipitation as a potential confounding influence on the time-series patterns, we obtained daily weather data from NOAA station USW00053152 in Santa Barbara (latitude: 34.4141, longitude: -119.8796) for 2020 to 2025 and summarized it to total annual precipitation. We then used a Pearson correlation to test the association between annual mean vegetation percent cover and annual precipitation. The correlation was not statistically significant (r = 0.82, p = 0.09), so we found no evidence that precipitation accounted for the observed cover trends; in any case, a correlation alone could not establish a causal relationship.

# Results

## Figure 1: Native Grass vs. Non-Native Vegetation Over Time

```{r}
#| label: timeseries-plot
#| message: false
#| echo: false
#| fig-width: 10
#| fig-height: 5
#| fig-cap: "Mean percent cover of native grasses and non-native vegetation along the seven Mesa (GL) transects at the North Campus Open Space (NCOS) from 2021 to 2025. Filled diamonds represent the grand mean across transects, and open circles represent individual transect means. Shaded ribbons show ±1 standard error. Dashed vertical lines indicate the Fall 2023 prescribed burn and the Fall 2024 rotational sheep grazing treatment events."

TS_plot <- ggplot() +
  # individual transect observations as open circles
  geom_jitter(data = grasses_transect,
             aes(x = year, y = mean_pc, color = cover_category),
             shape = 16, alpha = 0.4, size = 2, width = 0.1, height = 0.1) +
  # ribbon showing ±1 SE around the grand mean; low alpha to reduce visual noise
  geom_ribbon(data = grasses_summary,
              aes(x = year, ymin = grand_mean_pc - se, ymax = grand_mean_pc + se,
                  fill = cover_category), alpha = 0.1) +
  # line connecting the grand means across years
  geom_line(data = grasses_summary,
            aes(x = year, y = grand_mean_pc, color = cover_category)) +
  # grand mean as filled circle
  geom_point(data = grasses_summary,
             aes(x = year, y = grand_mean_pc, color = cover_category),
             shape = 18, size = 5) +
  # dashed line marking the prescribed burn intentional disturbance event
  geom_vline(xintercept = 2023.5, linetype = "dashed", color = "black") +
  # dashed line marking the rotational grazing intentional disturbance event
  geom_vline(xintercept = 2024.5, linetype = "dashed", color = "black") +
  # text label for burn event; hjust pushes label right of the line
  annotate("text", x = 2023.5, y = 10,
           label = "Prescribed Burn", hjust = -0.1, size = 3.5) +
  # text label for grazing event; hjust pushes label right of the line
  annotate("text", x = 2024.5, y = 10,
           label = "Sheep Grazing", hjust = -0.1, size = 3.5) +
  # setting color colors; colors must match to merge legends
  scale_color_manual(name = "Cover Type",
                   values = c("NATIVE COVER" = "#005AB5",
                              "NON-NATIVE COVER" = "#DC3220"),
                   labels = c("Native", "Non-native")) +
  # setting fill colors; colors must match to merge legends
  scale_fill_manual(name = "Cover Type",
                  values = c("NATIVE COVER" = "#005AB5",
                             "NON-NATIVE COVER" = "#DC3220"),
                  labels = c("Native", "Non-native")) +
  # labeling the axes
  labs(x = "Year",
       y = "Mean Percent Cover",
       color = "Cover Type",
       # consider dropping title since figure caption conveys the same information
       title = "Native Grasses vs. Non-Native Vegetation Over Time") +
  # adding black and white theme
  theme_bw() +
  # moving legend position to the top-left of figure panel
  theme(legend.position = c(0.15, 0.82),
        # transparent legend background so underlying data points are visible
        legend.background = element_rect(fill = "transparent"))

# displaying plot
TS_plot
```

Figure 1 answers study question 1 by tracing how native and non-native grass cover varied over time. From 2021 to 2023, the mean percent cover followed restoration as usual, with both native and non-native grasses increasing. Native grasses remained the more abundant group throughout the pre-burn period. Sub-question 1a is addressed by showing that this relationship reversed: native cover dropped sharply between 2023 and 2024 while non-native cover declined only slightly, so that non-native grasses edged above native grasses afterward. By 2025, the cover of both groups declined further, and the two nearly converged, with non-native grasses remaining marginally above native grasses.


## Figure 2: Mean Percent Cover Change Before and After the Burn

```{r}
#| label: plotting change in mean percent cover
#| message: false
#| echo: false
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "Change in mean percent cover of four key grassland species on the NCOS Mesa from 2023 (pre-burn) to 2024 (one year post-burn). Bars represent the mean percent-point change in cover across the seven Mesa transects for each species, and individual points show transect-level percent-point changes. The gray horizontal line marks no change; values above zero indicate increased cover and values below zero indicate decreased cover. Blue denotes the native species Stipa pulchra, and red denotes the invasive species Medicago polymorpha, Melilotus indicus, and Festuca perennis. Asterisks mark species with a statistically significant change in cover between the two years (paired t-test, p < 0.05). On average, S. pulchra cover declined and M. indicus cover increased, both significantly, whereas F. perennis declined and M. polymorpha changed little, neither significantly."

# create a ggplot with baselayer as the calculated percent point changes
percent_point_cover_change_plot <- ggplot(
  focal_change_summary_pp,
  # x axis = focal species, y axis = mean percent point change,
  # fill = native or invasive status
  aes(x = species_label, y = mean_percent_point_change, fill = status)
) +
  # bars show mean percent point change from 2023 to 2024
  geom_col(width = 0.6,
    alpha = 0.6
  ) +
  # jittered points show transect level percent point changes
  geom_jitter(
    data = focal_change_transect_pp,
    aes(x = species_label, y = percent_point_change, color = status),
    width = 0.08,
    height = 0,
    size = 2,
    inherit.aes = FALSE
  ) +
  # horizontal reference line at 0 showing no change between years
  geom_hline(
    yintercept = 0,
    color = "gray50"
  ) +
  # add rounded mean percent point change labels above or below columns
  geom_text(
    aes(
      label = round(mean_percent_point_change, 0),
      y = if_else(
        mean_percent_point_change >= 0,
        mean_percent_point_change + 2.9,
        mean_percent_point_change - 2.4
      )
    ),
    size = 3.5,
    color = "black",
    nudge_x = 0.15
  ) +
  # set fill colors for native and invasive bar categories
  scale_fill_manual(
    values = c(
      "Native" = "#005AB5",
      "Invasive" = "#DC3220"
    )
  ) +
  # set point colors to match bar fill colors
  scale_color_manual(
    values = c(
      "Native" = "#005AB5",
      "Invasive" = "#DC3220"
    ),
    guide = "none"
  ) +
  # add plot title, axis labels, and legend label
  labs(
    title = "Change in Mean % Cover of Key Species from Pre to Post-burn",
    x = "Species",
    y = "Change in Mean % Cover\n(percentage points)",
    fill = NULL
  ) +
  # adding bold and italic x labels
  scale_x_discrete(
    labels = c(
      "S. pulchra*"   = "***S. pulchra***\\*",
      "M. polymorpha" = "*M. polymorpha*",
      "M. indicus*"   = "***M. indicus***\\*",
      "F. perennis"   = "*F. perennis*"
    )
  ) +
  # non default theme
  theme_bw() +
  # consistent legend and axis text
  theme(
    legend.position = c(0.15, 0.82),
    legend.background = element_rect(fill = "transparent"),
    axis.text.x = ggtext::element_markdown()
  )

# display plot
percent_point_cover_change_plot
```

```{r}
#| label: calculating and storing t-test result in flextable
#| message: false
#| echo: false
#| warning: false
#| tbl-cap: "Paired t-test results comparing mean percent cover of four focal species before (2023) and after (2024) the prescribed burn along Mesa (GL) transects at NCOS (n = 7 transects). Significant p-values (p < 0.05) are bolded. Tests were conducted without correction for multiple comparisons."

# run paired t-test for each species and extract key statistics
focal_t_test_results <- focal_change_transect_pp |>
  # group by spp and spp label
  group_by(psoc, species_label) |>
  # calculate and store full t-test output as a list column
  summarize(
    # year_2023 and year 2024 are the mean percent cover values from respective years
    t_test = list(t.test(year_2023, year_2024, paired = TRUE))
  ) |>
  mutate(
    # extract t-statistic from each t-test result
    t_stat = map_dbl(t_test, ~ .x$statistic),
    # extract degrees of freedom from each t-test result
    df = map_dbl(t_test, ~ .x$parameter),
    # extract p-value from each t-test result
    p_value = map_dbl(t_test, ~ .x$p.value)
  ) |>
  # removing groups
  ungroup() |> 
  # drop psoc and t_test list column; keep only display columns
  select( species_label, t_stat, df, p_value)

# format results as a flextable for report output
focal_t_test_table <- focal_t_test_results |>
  flextable() |>
  # rename columns for readability
  set_header_labels(
    species_label = "Species",
    t_stat = "t-statistic",
    df = "df",
    p_value = "p-value"
  ) |>
  # round t-statistic and p-value to 3 decimal places
  colformat_double(j = c("t_stat", "p_value"), digits = 3) |>
  # italicize species names
  italic(j = "species_label") |>
  # bold significant p-values
  bold(i = ~ p_value < 0.05, j = "p_value") |>
  # auto fit column widths
  autofit()

# display table
focal_t_test_table
```

A paired t-test showed that two of our four focal species changed significantly in cover between 2023 and 2024 (Table 1). *Stipa pulchra*, the native restoration target, declined by roughly 21 percentage points between years (t = 4.66, df = 6, p = 0.003), the largest change of any species. This decline tracks the post-burn drop in aggregate native grass cover in Figure 1, since *Stipa pulchra* is the dominant native grass on the Mesa. In contrast, *Melilotus indicus* increased by roughly 14 percentage points between years (t = -3.51, df = 6, p = 0.013). The other two invasive species did not change significantly: *Festuca perennis* declined on average (about 14 percentage points) but had no statistical significance (p = 0.350), and *Medicago polymorpha* showed little average change but also had no significance (p = 0.716). 
Together, these results show that the burn did not affect species uniformly; responses were species-specific rather than predictable from nativity alone, with one native declining sharply, one invasive increasing sharply, and the remaining two invasives showing no significant change.

# Discussion

The NCOS 2024 Monitoring Report provides context for the cultural burn, allowing 2023 and 2024 to be interpreted as pre and post-burn comparison years. The decline in *Stipa pulchra* is important because it shows that the key native perennial grass indicator did not recover immediately one year after the burn. This alone does not determine the efficacy of intentional disturbance, but it identifies S. pulchra as a central species in continued monitoring efforts on the Mesa. Further, the report notes that *Melilotus indicus* emerged after fire at NCOS, a pattern reinforced by its post-burn proliferation in our results. This supports the idea that intentional disturbance created immediate conditions favorable for *M. indicus*, a non-native forb, and cannot simply be interpreted as suppressing invaders or promoting native grass recovery. The decline in *S. pulchra* and increase in *M. indicus* suggest that the Mesa’s ongoing post-burn trajectory must be evaluated through species-level comparisons, not only broad native versus non-native cover.

Previous grassland studies show that the effects of prescribed fire are species-specific and timescale-dependent. Grassland species are known to show varying degrees of reestablishment following fire (Berleman et al., 2016), which helps explain why the NCOS burn produced contrasting responses among focal species, and not a single community-wide recovery pattern. Additionally, the non-native persistence we observed after fire is consistent with research showing that one burn may not significantly reduce non-native abundance, especially when deeper persistent seedbanks remain largely unaffected by surface fire (Galloway et al., 2026). The decline in native grass cover should therefore be interpreted as an early response rather than a final restoration outcome. *Stipa pulchra* recovery may unfold over multiple years after prescribed burning, and restoration indicators may be idiosyncratic and do not always move steadily and predictably toward target conditions (Keeley et al., 2023; Matthews et al., 2009). The key question here is whether continued monitoring shows the community eventually moving towards native perennial grassland.

# Limitations

Several features of the data limit interpretation. Cover was estimated ocularly, so values are subjective and can vary between observers and years, and sampling was restricted to eight fixed transects, which may not represent the full spatial variation of the Mesa. With only eight transects and five survey years, both the t-tests and the precipitation correlation had limited statistical power.

Key confounding variables were also poorly resolved. Two of these potential confounding variables are precipitation and temperature, but the daily NOAA values (precipitation in inches, maximum temperature in Fahrenheit) had to be aggregated to annual summary values to match the yearly surveys, discarding the within-season timing that matters most and likely explaining why our precipitation correlation was not significant.

Temporal resolution is the final constraint. Surveys occur once per year in summer, when these grasses are largely dormant, so they miss the seasonal dynamics where much of the variation occurs. The annual cadence also cannot separate the Fall 2023 burn from the Fall 2024 grazing in the 2025 data, and a single post-burn year is too short to capture native perennial recovery.

# References
@nativeJamesgrass
@Hankins2013
@BERLEMAN2016423
@RAU2008169
@Carlsen2017
@MatthewsTraj
@KeeleyDecade
@galloway2026prescribed
@bartolome2019evaluating
@menke1992grazing
@rickard2024north
@nestor2025evaluating