---
title: "Final Paper"
author: "Lukas Lescano, Garrett Mellinger, Minh Tri Ngo"
format: pdf
bibliography: references.bib
---

```{r}
#| label: packages
#| message: false
#| warning: false
#| echo: false

# data wrangling and general use
library(tidyverse)
library(janitor)
library(here)

# plot enhancements
library(ggrepel)
library(cowplot)
library(ggtext)

# table formatting
library(flextable)

# saving tables
library(webshot2)
```

```{r}
#| label: data
#| message: false
#| echo: false

# read in veg data
veg <- read_csv(here("data", "raw", "veg.csv"))

# read in metadata
metadata <- read_csv(here("data", "raw", "vp_veg_metadata.csv"))
```

```{r}
#| label: adding target species list
#| message: false
#| echo: false

# all grass species (Poaceae) at NCOS mesa, both native and non-native
poaceae_species <- c(
  "Bromus_hordeaceus",
  "Festuca_perennis",
  "Stipa_pulchra",
  "Polypogon_monspeliensis",
  "Brachypodium_distachyon",
  "Festuca_myuros",
  "Avena_barbata",
  "Hordeum_marinum",
  "Avena_fatua",
  "Parapholis_incurva",
  "Bromus_diandrus",
  "Distichlis_spicata",
  "Bromus_catharticus",
  "Hordeum_murinum",
  "Hordeum_brachyantherum_brachy.",
  "Polypogon_viridis",
  "Cortaderia_selloana",
  "Festuca_bromoides",
  "Cynodon_dactylon",
  "Poa_annua",
  "Festuca_microstachys",
  "Crypsis_schoenoides",
  "Polypogon_interruptus",
  "Hordeum_brachyantherum",
  "Bromus_madritensis",
  "Elymus_triticoides"
)

# native forb species to exclude from grass analysis
native_forbs <- c(
  "Plantago_erecta",
  "Sisyrinchium_bellum",
  "Sidalcea_malviflora",
  "Eryngium_vaseyi",
  "Acmispon_maritimus",
  "Acmispon_wrangelianus",
  "Madia_sativa",
  "Psilocarphus_tennulus",
  "Psilocarphus_brevissimus",
  "Spergularia_macrotheca",
  "Centromadia_parryi_australis",
  "Lepidium_nitidum",
  "Stebbinsoseris_heterocarpa",
  "Laennecia_coulteri"
)

# non-vegetation cover and non-target species to exclude from analysis
exclude_other <- c(
  "Bare Ground",
  "Natural Thatch",
  "unlisted",
  "moss",
  "boulder_rock",
  "biocrust",
  "Baccharis_pilularis",
  "Juncus_bufonius",
  "Juncus_phaeocephalus",
  "Eleocharis_macrostachya",
  "Eleocharis_acicularis",
  "Eleocharis_palustris"
)
```


```{r}
#| label: creating species lookup
#| message: false
#| echo: false

# creating data frame from veg
species_lookup <- veg |>
  # keep only GL transects at ncos
  filter(str_detect(transect_name, "GL"), site == "ncos") |>
  # remove unlisted since it maps to multiple categories
  filter(psoc != "unlisted") |>
  # return one unique row per species-category combination
  distinct(psoc, cover_category) |>
  # sort alphabetically by species for readability
  arrange(psoc)
```


```{r}
#| label: creating explicit0 dataframe
#| message: false
#| echo: false

# create a new object from veg
gl_explicit0 <- veg |> 
  # filter all occurrences where transect_name contains VP
  filter(str_detect(transect_name, "GL"),
         site == "ncos") |> 
  # not including cover_category because we will join it back after complete()
  select(-cover_category) |> 
  # complete all possible combinations of year, pool, psoc and distance
  complete(year, transect_name, psoc, transect_distance,
           # fill values of percent cover with 0
           fill = list(percent_cover = 0)) |>
  # group by year, pool, and psoc
  group_by(year, transect_name, psoc) |>
  # calculate total percent cover across all quads
  summarize(sum_pc = sum(percent_cover, na.rm = TRUE)) |>
  # ungroup the data frame
  ungroup() |>
  # create a new column called year_pool from year and transect name
  unite("year_pool", year, transect_name, remove = TRUE) |> 
  # join with metadata data frame
  left_join(metadata, by = "year_pool") |> 
  # calculate mean percent cover from total percent cover/number of quadrats
  mutate(mean_pc = sum_pc/num_quad) |> 
  # re-attach cover_category from lookup using psoc as key
  left_join(species_lookup, by = "psoc") |> 
  # excluding "GL-11" because it was unburned
  filter(transect_name != "GL-11")
```

```{r}
#| label: calculating change in mean percent cover
#| message: false
#| echo: false
#| warning: false

## Change in percentage points of focal species percent cover
## 2023 = pre-burn
## 2024 = post-burn

# define focal species with display labels and native/invasive status
focal_species <- tibble(
  psoc = c(
    "Stipa_pulchra",
    "Medicago_polymorpha",
    "Melilotus_indicus",
    "Festuca_perennis"
  ),
  species_label = c(
    "S. pulchra*",
    "M. polymorpha",
    "M. indicus*",
    "F. perennis"
  ),
  status = c(
    "Native",
    "Invasive",
    "Invasive",
    "Invasive"
  )
)

# filter gl_explicit0 to focal species and pre/post burn years only
focal_cover <- gl_explicit0 |>
  filter(
    year %in% c(2023, 2024),
    psoc %in% focal_species$psoc
  ) |>
  # keep only columns needed for change calculation
  select(year, transect_name, psoc, mean_pc) |>
  # attach species labels and status from focal_species
  left_join(focal_species, by = "psoc")

# calculate per-transect percentage point change from 2023 to 2024
focal_change_transect_pp <- focal_cover |>
  # pivot to wide format so 2023 and 2024 are separate columns
  pivot_wider(
    names_from = year,
    values_from = mean_pc,
    names_prefix = "year_"
  ) |>
  # calculate difference in mean percent cover between years
  mutate(
    percent_point_change = year_2024 - year_2023
  ) |>
  # remove transects missing data for either year
  filter(!is.na(percent_point_change))

# aggregate to mean percentage point change per species across transects
focal_change_summary_pp <- focal_change_transect_pp |>
  # group by spp, spp label, and status
  group_by(psoc, species_label, status) |>
  # calculate mean percent point change
  summarize(
    mean_percent_point_change = mean(percent_point_change, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) |>
  # set factor levels to control species order in plots
  mutate(
    species_label = factor(
      species_label,
      levels = c(
        "S. pulchra*",
        "M. polymorpha",
        "M. indicus*",
        "F. perennis"
      )
    ),
    # set factor levels to control fill order in legend
    status = factor(status, levels = c("Native", "Invasive"))
  )
```

```{r}
#| label: plot 1 creating grasses dataframe
#| message: false
#| echo: false

# creating grasses data frame from gl_explicit0
grasses <- gl_explicit0 |> 
  # filtering to include poaceae grasses species
  filter(psoc %in% poaceae_species,
         # filtering to include only native and non-native cover categories
         cover_category %in% c("NATIVE COVER", "NON-NATIVE COVER"))
```

```{r}
#| label: plot 1 creating grasses_summary for time series plot
#| message: false
#| echo: false

grasses_summary <- grasses |> 
  # grouping by year and cover category
  group_by(year, cover_category) |> 
  # calculating average per transect mean_pc across all transects and species within each group
  summarise(grand_mean_pc = mean(mean_pc, na.rm = TRUE),
            se = sd(mean_pc, na.rm = TRUE) / sqrt(n())) |> 
  # remove grouping
  ungroup()
```

```{r}
#| label: plot 1 creating grasses transect data frame
#| message: false
#| echo: false

# creating new object from grasses
grasses_transect <- grasses |>
  # grouping by year, transect, and conver category
  group_by(year, transect_name, cover_category) |>
  # calculating mean pc
  summarize(mean_pc = mean(mean_pc, na.rm = TRUE)) |>
  # removing groups
  ungroup()
```


# Introduction

# Methods
  
Analysis of native grasses and non-native vegetation was focused on the years 2021 to 2025, with particular interest in the years 2023 and 2024. The survey month for grasslands is in July.  Species of interest were chosen based on the overlap of target species from previous studies of sheep-grazing and burn disturbance. Within the 16.8 acres of perennial grassland (Mesa), permanent transects with one-square-meter quadrats were used to monitor specifically short or low-growing vegetation (Rickard, 2024). Quadrats alternated between left and right of the transect line every 3 meters, with transects being 30 meters long. The first quadrat is centered left of the starting point of each transect, giving 11 quadrats for a 30-meter transect, for a total of 8 transects, with 88 quadrats.  
  
Focus was placed on the 2023 cultural burn and its effects. Plots in grassland regions were surveyed before and after burning in July 2023 and May 2024, respectively. In the 14+ acres of perennial grasslands, NCOS staff used 65 1-meter by 1-meter experimental plots that were placed within the burn area, with an additional 10 taken outside the burn area to represent unburned controls (Rickard, 2024). Of these 65 plots, 32 were grass plots chosen based on the dominant grass species at each plot that contained the target species Sipta pulchra and Festuca perennis. 
 
For burns, percent cover data was collected by ocularly identifying each species within the plot and estimating its absolute percent cover of the plot. For quadrat transects, they are divided into 100 ten-centimeter squares and percent cover is estimated for each species in the quadrat. Additional figure inspiration was taken from the percent biomass change in sheep-grazing, using the percent change in cover for our analysis instead. 

To address our third question, whether individual focal species changed in cover before and after the burn, we used paired t-tests comparing mean percent cover in 2023 (pre-burn) and 2024 (post-burn) for each of the four focal species: Stipa pulchra, Medicago polymorpha, Melilotus indicus, and Festuca perennis. Observations were paired by transect (n = 7), so each test evaluated whether the per-transect change in cover between the two years differed from zero. We assessed significance at α = 0.05 and did not apply a multiple-comparison correction, given the small number of pre-specified focal species. All tests were conducted in R using the t.test function.

To explore precipitation as a potential confounding influence on the time-series patterns, we obtained daily weather data from NOAA station USW00053152 in Santa Barbara (latitude: 34.4141, longitude: -119.8796) for 2020 to 2025 and summarized it to annual precipitation. We then used a Pearson correlation to test the association between annual mean grass percent cover and annual precipitation. The correlation was not statistically significant (r = 0.82, p = 0.09), so we found no evidence that precipitation accounted for the observed cover trends; in any case, a correlation alone could not establish a causal relationship.


# Results

## Figure 1: Native Grass vs. Non-Native Vegetation Over Time

```{r}
#| label: timeseries-plot
#| message: false
#| echo: false
#| fig-width: 10
#| fig-height: 5
#| fig-cap: "Mean percent cover of native grasses and non-native vegetation along the seven Mesa (GL) transects at the North Campus Open Space (NCOS) from 2021 to 2025. Filled diamonds represent the grand mean across transects, and open circles represent individual transect means. Shaded ribbons show ±1 standard error. Dashed vertical lines indicate the Fall 2023 prescribed burn and the Fall 2024 rotational sheep grazing treatment events."

TS_plot <- ggplot() +
  # individual transect observations as open circles
  geom_jitter(data = grasses_transect,
             aes(x = year, y = mean_pc, color = cover_category),
             shape = 16, alpha = 0.4, size = 2, width = 0.1, height = 0.1) +
  # ribbon showing ±1 SE around the grand mean; low alpha to reduce visual noise
  geom_ribbon(data = grasses_summary,
              aes(x = year, ymin = grand_mean_pc - se, ymax = grand_mean_pc + se,
                  fill = cover_category), alpha = 0.1) +
  # line connecting the grand means across years
  geom_line(data = grasses_summary,
            aes(x = year, y = grand_mean_pc, color = cover_category)) +
  # grand mean as filled circle
  geom_point(data = grasses_summary,
             aes(x = year, y = grand_mean_pc, color = cover_category),
             shape = 18, size = 5) +
  # dashed line marking the prescribed burn intentional disturbance event
  geom_vline(xintercept = 2023.5, linetype = "dashed", color = "black") +
  # dashed line marking the rotational grazing intentional disturbance event
  geom_vline(xintercept = 2024.5, linetype = "dashed", color = "black") +
  # text label for burn event; hjust pushes label right of the line
  annotate("text", x = 2023.5, y = 10,
           label = "Prescribed Burn", hjust = -0.1, size = 3.5) +
  # text label for grazing event; hjust pushes label right of the line
  annotate("text", x = 2024.5, y = 10,
           label = "Sheep Grazing", hjust = -0.1, size = 3.5) +
  # setting color colors; colors must match to merge legends
  scale_color_manual(name = "Cover Type",
                   values = c("NATIVE COVER" = "#005AB5",
                              "NON-NATIVE COVER" = "#DC3220"),
                   labels = c("Native", "Non-native")) +
  # setting fill colors; colors must match to merge legends
  scale_fill_manual(name = "Cover Type",
                  values = c("NATIVE COVER" = "#005AB5",
                             "NON-NATIVE COVER" = "#DC3220"),
                  labels = c("Native", "Non-native")) +
  # labeling the axes
  labs(x = "Year",
       y = "Mean Percent Cover",
       color = "Cover Type",
       # consider dropping title since figure caption conveys the same information
       title = "Native Grasses vs. Non-Native Vegetation Over Time") +
  # adding black and white theme
  theme_bw() +
  # moving legend position to the top-left of figure panel
  theme(legend.position = c(0.15, 0.82),
        # transparent legend background so underlying data points are visible
        legend.background = element_rect(fill = "transparent"))

# displaying plot
TS_plot
```

## Figure 2: Mean Percent Cover Change Before and After the Burn

```{r}
#| label: plotting change in mean percent cover
#| message: false
#| echo: false
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "Change in mean percent cover of four key grassland species on the NCOS Mesa from 2023 (pre-burn) to 2024 (one year post-burn). Bars represent the mean percent-point change in cover across the seven Mesa transects for each species, and individual points show transect-level percent-point changes. The gray horizontal line marks no change; values above zero indicate increased cover and values below zero indicate decreased cover. Blue denotes the native species Stipa pulchra, and red denotes the invasive species Medicago polymorpha, Melilotus indicus, and Festuca perennis. Asterisks mark species with a statistically significant change in cover between the two years (paired t-test, p < 0.05). On average, S. pulchra cover declined and M. indicus cover increased, both significantly, whereas F. perennis declined and M. polymorpha changed little, neither significantly."

# create a ggplot with baselayer as the calculated percent point changes
percent_point_cover_change_plot <- ggplot(
  focal_change_summary_pp,
  # x axis = focal species, y axis = mean percent point change,
  # fill = native or invasive status
  aes(x = species_label, y = mean_percent_point_change, fill = status)
) +
  # bars show mean percent point change from 2023 to 2024
  geom_col(width = 0.6,
    alpha = 0.6
  ) +
  # jittered points show transect level percent point changes
  geom_jitter(
    data = focal_change_transect_pp,
    aes(x = species_label, y = percent_point_change, color = status),
    width = 0.08,
    height = 0,
    size = 2,
    inherit.aes = FALSE
  ) +
  # horizontal reference line at 0 showing no change between years
  geom_hline(
    yintercept = 0,
    color = "gray50"
  ) +
  # add rounded mean percent point change labels above or below columns
  geom_text(
    aes(
      label = round(mean_percent_point_change, 0),
      y = if_else(
        mean_percent_point_change >= 0,
        mean_percent_point_change + 2.9,
        mean_percent_point_change - 2.4
      )
    ),
    size = 3.5,
    color = "black",
    nudge_x = 0.15
  ) +
  # set fill colors for native and invasive bar categories
  scale_fill_manual(
    values = c(
      "Native" = "#005AB5",
      "Invasive" = "#DC3220"
    )
  ) +
  # set point colors to match bar fill colors
  scale_color_manual(
    values = c(
      "Native" = "#005AB5",
      "Invasive" = "#DC3220"
    ),
    guide = "none"
  ) +
  # add plot title, axis labels, and legend label
  labs(
    title = "Change in Mean % Cover of Key Species from Pre to Post-burn",
    x = "Species",
    y = "Change in Mean % Cover\n(percentage points)",
    fill = NULL
  ) +
  # adding bold and italic x labels
  scale_x_discrete(
    labels = c(
      "S. pulchra*"   = "***S. pulchra***\\*",
      "M. polymorpha" = "*M. polymorpha*",
      "M. indicus*"   = "***M. indicus***\\*",
      "F. perennis"   = "*F. perennis*"
    )
  ) +
  # non default theme
  theme_bw() +
  # consistent legend and axis text
  theme(
    legend.position = c(0.15, 0.82),
    legend.background = element_rect(fill = "transparent"),
    axis.text.x = ggtext::element_markdown()
  )

# display plot
percent_point_cover_change_plot
```

# Discussion

# References
@nativeJamesgrass
@Hankins2013
@BERLEMAN2016423
@RAU2008169
@Carlsen2017
@MatthewsTraj
@KeeleyDecade
@galloway2026prescribed
@bartolome2019evaluating
@menke1992grazing
@rickard2024north
@nestor2025evaluating