---
title: "Timeline Check in"
author: "Lukas Lescano, Garrett Mellinger, Minh Tri Ngo"
date: "2026-05-22"
format: pdf
message: false
warning: false
---

```{r}
#| label: packages
#| message: false
#| echo: false

# data wrangling and general use
library(tidyverse)
library(janitor)
library(here)

# plot enhancements
library(ggrepel)
library(cowplot)
library(ggtext)

# table formatting
library(flextable)

# saving tables
library(webshot2)
```


```{r}
#| label: data
#| message: false
#| echo: false

# read in veg data
veg <- read_csv(here("data", "raw", "veg.csv"))

# read in metadata
metadata <- read_csv(here("data", "raw", "vp_veg_metadata.csv"))
```


```{r}
#| label: adding target species list
#| message: false
#| echo: false

# all grass species (Poaceae) at NCOS mesa, both native and non-native
poaceae_species <- c(
  "Bromus_hordeaceus",
  "Festuca_perennis",
  "Stipa_pulchra",
  "Polypogon_monspeliensis",
  "Brachypodium_distachyon",
  "Festuca_myuros",
  "Avena_barbata",
  "Hordeum_marinum",
  "Avena_fatua",
  "Parapholis_incurva",
  "Bromus_diandrus",
  "Distichlis_spicata",
  "Bromus_catharticus",
  "Hordeum_murinum",
  "Hordeum_brachyantherum_brachy.",
  "Polypogon_viridis",
  "Cortaderia_selloana",
  "Festuca_bromoides",
  "Cynodon_dactylon",
  "Poa_annua",
  "Festuca_microstachys",
  "Crypsis_schoenoides",
  "Polypogon_interruptus",
  "Hordeum_brachyantherum",
  "Bromus_madritensis",
  "Elymus_triticoides"
)

# native forb species to exclude from grass analysis
native_forbs <- c(
  "Plantago_erecta",
  "Sisyrinchium_bellum",
  "Sidalcea_malviflora",
  "Eryngium_vaseyi",
  "Acmispon_maritimus",
  "Acmispon_wrangelianus",
  "Madia_sativa",
  "Psilocarphus_tennulus",
  "Psilocarphus_brevissimus",
  "Spergularia_macrotheca",
  "Centromadia_parryi_australis",
  "Lepidium_nitidum",
  "Stebbinsoseris_heterocarpa",
  "Laennecia_coulteri"
)

# non-vegetation cover and non-target species to exclude from analysis
exclude_other <- c(
  "Bare Ground",
  "Natural Thatch",
  "unlisted",
  "moss",
  "boulder_rock",
  "biocrust",
  "Baccharis_pilularis",
  "Juncus_bufonius",
  "Juncus_phaeocephalus",
  "Eleocharis_macrostachya",
  "Eleocharis_acicularis",
  "Eleocharis_palustris"
)
```


```{r}
#| label: creating species lookup
#| message: false
#| echo: false

# creating data frame from veg
species_lookup <- veg |>
  # keep only GL transects at ncos
  filter(str_detect(transect_name, "GL"), site == "ncos") |>
  # remove unlisted since it maps to multiple categories
  filter(psoc != "unlisted") |>
  # return one unique row per species-category combination
  distinct(psoc, cover_category) |>
  # sort alphabetically by species for readability
  arrange(psoc)
```


```{r}
#| label: creating explicit0 dataframe
#| message: false
#| echo: false

# create a new object from veg
gl_explicit0 <- veg |> 
  # filter all occurrences where transect_name contains VP
  filter(str_detect(transect_name, "GL"),
         site == "ncos") |> 
  # not including cover_category because we will join it back after complete()
  select(-cover_category) |> 
  # complete all possible combinations of year, pool, psoc and distance
  complete(year, transect_name, psoc, transect_distance,
           # fill values of percent cover with 0
           fill = list(percent_cover = 0)) |>
  # group by year, pool, and psoc
  group_by(year, transect_name, psoc) |>
  # calculate total percent cover across all quads
  summarize(sum_pc = sum(percent_cover, na.rm = TRUE)) |>
  # ungroup the data frame
  ungroup() |>
  # create a new column called year_pool from year and transect name
  unite("year_pool", year, transect_name, remove = TRUE) |> 
  # join with metadata data frame
  left_join(metadata, by = "year_pool") |> 
  # calculate mean percent cover from total percent cover/number of quadrats
  mutate(mean_pc = sum_pc/num_quad) |> 
  # re-attach cover_category from lookup using psoc as key
  left_join(species_lookup, by = "psoc") |> 
  # excluding "GL-11" because it was unburned
  filter(transect_name != "GL-11")
```

# Ideas for analysis

## Questions

- How does the percent cover of native grasses vary over time compared to the percent cover of exotic vegetation on the NCOS Mesa Perennial Grasslands?
  - How does intentional disturbance influence this relationship?
  - How do key individual species, including *Stipa pulchra*, *Medicago polymorpha*, *Melilotus indicus*, and *Festuca perennis*, vary in percent cover before and after intentional disturbance?

## Statistical analysis
- statistical analysis to address at least one question with at least one analysis coded and run

```{r}
#| label: calculating change in mean percent cover
#| message: false
#| echo: false
#| warning: false

## Change in percentage points of focal species percent cover
## 2023 = pre-burn
## 2024 = post-burn

# define focal species with display labels and native/invasive status
focal_species <- tibble(
  psoc = c(
    "Stipa_pulchra",
    "Medicago_polymorpha",
    "Melilotus_indicus",
    "Festuca_perennis"
  ),
  species_label = c(
    "S. pulchra",
    "M. polymorpha",
    "M. indicus",
    "F. perennis"
  ),
  status = c(
    "Native",
    "Invasive",
    "Invasive",
    "Invasive"
  )
)

# filter gl_explicit0 to focal species and pre/post burn years only
focal_cover <- gl_explicit0 |>
  filter(
    year %in% c(2023, 2024),
    psoc %in% focal_species$psoc
  ) |>
  # keep only columns needed for change calculation
  select(year, transect_name, psoc, mean_pc) |>
  # attach species labels and status from focal_species
  left_join(focal_species, by = "psoc")

# calculate per-transect percentage point change from 2023 to 2024
focal_change_transect_pp <- focal_cover |>
  # pivot to wide format so 2023 and 2024 are separate columns
  pivot_wider(
    names_from = year,
    values_from = mean_pc,
    names_prefix = "year_"
  ) |>
  # calculate difference in mean percent cover between years
  mutate(
    percent_point_change = year_2024 - year_2023
  ) |>
  # remove transects missing data for either year
  filter(!is.na(percent_point_change))

# aggregate to mean percentage point change per species across transects
focal_change_summary_pp <- focal_change_transect_pp |>
  # group by spp, spp label, and status
  group_by(psoc, species_label, status) |>
  # calculate mean percent point change
  summarize(
    mean_percent_point_change = mean(percent_point_change, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) |>
  # set factor levels to control species order in plots
  mutate(
    species_label = factor(
      species_label,
      levels = c(
        "S. pulchra",
        "M. polymorpha",
        "M. indicus",
        "F. perennis"
      )
    ),
    # set factor levels to control fill order in legend
    status = factor(status, levels = c("Native", "Invasive"))
  )
```

```{r}
#| label: calculating and storing t-test result in flextable
#| message: false
#| echo: true
#| warning: false
#| tbl-cap: "Paired t-test results comparing mean percent cover of four focal species before (2023) and after (2024) the prescribed burn along Mesa (GL) transects at NCOS (n = 7 transects). Significant p-values (p < 0.05) are bolded. Tests were conducted without correction for multiple comparisons."

# run paired t-test for each species and extract key statistics
focal_t_test_results <- focal_change_transect_pp |>
  # group by spp and spp label
  group_by(psoc, species_label) |>
  # calculate and store full t-test output as a list column
  summarize(
    # year_2023 and year 2024 are the mean percent cover values from respective years
    t_test = list(t.test(year_2023, year_2024, paired = TRUE))
  ) |>
  mutate(
    # extract t-statistic from each t-test result
    t_stat = map_dbl(t_test, ~ .x$statistic),
    # extract degrees of freedom from each t-test result
    df = map_dbl(t_test, ~ .x$parameter),
    # extract p-value from each t-test result
    p_value = map_dbl(t_test, ~ .x$p.value)
  ) |>
  # removing groups
  ungroup() |> 
  # drop psoc and t_test list column; keep only display columns
  select( species_label, t_stat, df, p_value)

# format results as a flextable for report output
focal_t_test_table <- focal_t_test_results |>
  flextable() |>
  # rename columns for readability
  set_header_labels(
    species_label = "Species",
    t_stat = "t-statistic",
    df = "df",
    p_value = "p-value"
  ) |>
  # round t-statistic and p-value to 3 decimal places
  colformat_double(j = c("t_stat", "p_value"), digits = 3) |>
  # italicize species names
  italic(j = "species_label") |>
  # bold significant p-values
  bold(i = ~ p_value < 0.05, j = "p_value") |>
  # auto fit column widths
  autofit()

focal_t_test_table <- bg(focal_t_test_table, bg = "white", part = "all")

# display table
focal_t_test_table
```

# Background

We assess whether the 2023 cultural burn is meeting its restoration goals of promoting native grassland species. This study may help to inform future restoration management decisions regarding intentional disturbance at NCOS and could have implications which extend to the management or ecological understanding of other coastal grassland systems. These questions are relevant because Santa Barbara coastal grasslands are fire adapted (Saglimbeni, 2024, pp. 1-2), but are now heavily affected by invasive species. Studying the NCOS burn helps show whether cultural burning can support native species like Stipa pulchra while reducing nonnative grassland invaders.


# Visualizations

## Plot 1: Native and exotic percent cover over time

```{r}
#| label: creating grasses dataframe
#| message: false
#| echo: false

# creating grasses data frame from gl_explicit0
grasses <- gl_explicit0 |> 
  # filtering to include poaceae grasses species
  filter(psoc %in% poaceae_species,
         # filtering to include only native and non-native cover categories
         cover_category %in% c("NATIVE COVER", "NON-NATIVE COVER"))
```

```{r}
#| label: creating grasses_summary for time series plot
#| message: false
#| echo: false

grasses_summary <- grasses |> 
  # grouping by year and cover category
  group_by(year, cover_category) |> 
  # calculating average per transect mean_pc across all transects and species within each group
  summarise(grand_mean_pc = mean(mean_pc, na.rm = TRUE),
            se = sd(mean_pc, na.rm = TRUE) / sqrt(n())) |> 
  # remove grouping
  ungroup()
```

```{r}
#| label: creating grasses transect data frame
#| message: false
#| echo: false

# creating new object from grasses
grasses_transect <- grasses |>
  # grouping by year, transect, and conver category
  group_by(year, transect_name, cover_category) |>
  # calculating mean pc
  summarize(mean_pc = mean(mean_pc, na.rm = TRUE)) |>
  # removing groups
  ungroup()
```

```{r}
#| label: timeseries-plot
#| message: false
#| echo: false
#| fig-width: 10
#| fig-height: 5
#| fig-cap: "Mean percent cover of native and non-native grasses along the Mesa (GL) transects at the North Campus Open Space (NCOS) from 2021 to 2025. Filled circles represent the grand mean across transects and open circles represent individual transect means. Shaded ribbons show ±1 standard error. Dashed vertical lines indicate the prescribed burn (Fall 2023) and rotational sheep grazing (Fall 2024) treatment events."

TS_plot <- ggplot() +
  # individual transect observations as open circles
  geom_jitter(data = grasses_transect,
             aes(x = year, y = mean_pc, color = cover_category),
             shape = 16, alpha = 0.4, size = 2, width = 0.1, height = 0.1) +
  # ribbon showing ±1 SE around the grand mean; low alpha to reduce visual noise
  geom_ribbon(data = grasses_summary,
              aes(x = year, ymin = grand_mean_pc - se, ymax = grand_mean_pc + se,
                  fill = cover_category), alpha = 0.1) +
  # line connecting the grand means across years
  geom_line(data = grasses_summary,
            aes(x = year, y = grand_mean_pc, color = cover_category)) +
  # grand mean as filled circle
  geom_point(data = grasses_summary,
             aes(x = year, y = grand_mean_pc, color = cover_category),
             shape = 18, size = 5) +
  # dashed line marking the prescribed burn intentional disturbance event
  geom_vline(xintercept = 2023.5, linetype = "dashed", color = "black") +
  # dashed line marking the rotational grazing intentional disturbance event
  geom_vline(xintercept = 2024.5, linetype = "dashed", color = "black") +
  # text label for burn event; hjust pushes label right of the line
  annotate("text", x = 2023.5, y = 10,
           label = "Prescribed Burn", hjust = -0.1, size = 3.5) +
  # text label for grazing event; hjust pushes label right of the line
  annotate("text", x = 2024.5, y = 10,
           label = "Sheep Grazing", hjust = -0.1, size = 3.5) +
  # setting color colors; colors must match to merge legends
  scale_color_manual(name = "Cover Type",
                   values = c("NATIVE COVER" = "#005AB5",
                              "NON-NATIVE COVER" = "#DC3220"),
                   labels = c("Native", "Non-native")) +
  # setting fill colors; colors must match to merge legends
  scale_fill_manual(name = "Cover Type",
                  values = c("NATIVE COVER" = "#005AB5",
                             "NON-NATIVE COVER" = "#DC3220"),
                  labels = c("Native", "Non-native")) +
  # labeling the axes
  labs(x = "Year",
       y = "Mean Percent Cover",
       color = "Cover Type",
       # consider dropping title since figure caption conveys the same information
       title = "Native vs. Non-Native Grass Cover Over Time") +
  # moving legend position to the top-left of figure panel
  theme(legend.position = c(0.15, 0.82),
        # transparent legend background so underlying data points are visible
        legend.background = element_rect(fill = "transparent")) +
  # adding black and white theme
  theme_bw()

# displaying plot
TS_plot
```

## Plot 2: Sweet and bur clover abundance over time

```{r}
#| label: plot 2 wrangling
#| message: false
#| echo: false

# creating new object from gl_explicit0
clover <- gl_explicit0 |>
  # filter to clover species of interest
  filter(psoc %in% c("Melilotus_indicus", "Medicago_polymorpha"))
```

```{r}
#| label: creating clover summary data frame
#| message: false
#| echo: false

# creating new object from clover
clover_summary <- clover |>
  # grouping by year and psoc
  group_by(year, psoc) |>
  # calculating grand mean percent cover
  summarize(grand_mean_pc = mean(mean_pc, na.rm = TRUE),
            # calculating standard error
            se = sd(mean_pc, na.rm = TRUE) / sqrt(n())) |>
  # removing groups
  ungroup()
```

```{r}
#| label: creating clover plot
#| message: false
#| echo: false
#| fig-width: 10
#| fig-height: 5
#| fig-cap: "Mean percent cover of bur clover (*M. polymorpha*) and sweet clover (*M. indicus*) along the Mesa (GL) transects at the North Campus Open Space (NCOS) from 2021 to 2025. Filled circles represent the grand mean across transects and open circles represent individual transect means. Shaded ribbons show ±1 standard error. Dashed vertical lines indicate the prescribed burn (Fall 2023) and rotational sheep grazing (Fall 2024) treatment events. Field observations suggested *M. polymorpha* increased following the prescribed burn and *M. indicus* following sheep grazing, a pattern reflected in the monitoring data."

# creating new plot
clover_plot <- ggplot() +
  # individual transect observations as open circles
  geom_jitter(data = clover,
              aes(x = year, y = mean_pc, color = psoc),
              shape = 16, alpha = 0.4, size = 2,
              width = 0.1, height = 0.1) +
  # ribbon showing standard error around the grand mean
  geom_ribbon(data = clover_summary,
              aes(x = year, ymin = grand_mean_pc - se, ymax = grand_mean_pc + se,
                  fill = psoc), alpha = 0.1) +
  # line connecting grand means across years
  geom_line(data = clover_summary,
            aes(x = year, y = grand_mean_pc, color = psoc)) +
  # grand mean as filled circle
  geom_point(data = clover_summary,
             aes(x = year, y = grand_mean_pc, color = psoc),
             shape = 18, size = 5) +
  # dashed line marking prescribed burn between 2023 and 2024 surveys
  geom_vline(xintercept = 2023.5, linetype = "dashed", color = "black") +
  # dashed line marking rotational sheep grazing between 2024 and 2025 surveys
  geom_vline(xintercept = 2024.5, linetype = "dashed", color = "black") +
  # text label for burn event
  annotate("text", x = 2023.5, y = 42,
           label = "Prescribed Burn", hjust = -0.1, size = 3.5) +
  # text label for grazing event
  annotate("text", x = 2024.5, y = 42,
           label = "Sheep Grazing", hjust = -0.1, size = 3.5) +
  # setting color colors; colors must match to merge legends
  scale_color_manual(name = "Species",
                   values = c("Melilotus_indicus" = "#E66100",
                              "Medicago_polymorpha" = "#5D3A9B"),
                   labels = c("<i>M. indicus</i> (sweet clover)", 
                              "<i>M. polymorpha</i> (bur clover)")) +
  # setting fill colors; colors must match to merge legends
  scale_fill_manual(name = "Species",
                  values = c("Melilotus_indicus" = "#E66100",
                             "Medicago_polymorpha" = "#5D3A9B"),
                  labels = c("<i>M. indicus</i> (sweet clover)", 
                             "<i>M. polymorpha</i> (bur clover)")) +
  # labeling axes
  labs(x = "Year",
       y = "Mean Percent Cover",
       title = "Sweet and Bur Clover Cover Over Time") +
  # enable HTML markdown to render in legend text to display the italic species names
  theme(legend.text = element_markdown(),
        # moving legend position to the top-left of figure panel
        legend.position = c(0.15, 0.82),
        # transparent legend background so underlying data points are visible
        legend.background = element_rect(fill = "transparent")) +
  # adding black and white theme
  theme_bw()

# displaying plot
clover_plot
```

## Plot 3: Change in mean percent cover of key species after disturbance

```{r}
#| label: plotting change in mean percent cover
#| message: false
#| echo: false
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "Change in mean percent cover of four key grassland species on the NCOS Mesa from 2023, pre-burn, to 2024, one year post-burn. Bars represent the average percent-point change in cover across transects for each species, while individual points show transect-level percent-point changes; the gray horizontal line represents no change, with values above zero indicating increased cover and values below zero indicating decreased cover. Blue indicates the native species, Stipa pulchra, and red indicates invasive species: Medicago polymorpha, Melilotus indicus, and Festuca perennis. Overall, *S. pulchra* and *F. perennis* decreased after the burn, *M. indicus* increased, and *M. polymorpha* showed little average change. This figure connects to our main question by showing that intentional disturbance did not affect all native and invasive species in the same way."

# create a ggplot with baselayer as the calculated percent point changes
percent_point_cover_change_plot <- ggplot(
  focal_change_summary_pp,
  # x axis = focal species, y axis = mean percent point change,
  # fill = native or invasive status
  aes(x = species_label, y = mean_percent_point_change, fill = status)
) +
  # bars show mean percent point change from 2023 to 2024
  geom_col(width = 0.6) +
  # jittered points show transect level percent point changes
  geom_jitter(
    data = focal_change_transect_pp,
    aes(x = species_label, y = percent_point_change, color = status),
    width = 0.08,
    height = 0,
    size = 2,
    alpha = 0.6,
    inherit.aes = FALSE
  ) +
  # horizontal reference line at 0 showing no change between years
  geom_hline(
    yintercept = 0,
    color = "gray50"
  ) +
  # add rounded mean percent point change labels above or below columns
  geom_text(
    aes(
      label = round(mean_percent_point_change, 0),
      y = if_else(
        mean_percent_point_change >= 0,
        mean_percent_point_change + 2.9,
        mean_percent_point_change - 2.4
      )
    ),
    size = 5,
    color = "black"
  ) +
  # set fill colors for native and invasive bar categories
  scale_fill_manual(
    values = c(
      "Native" = "#005AB5",
      "Invasive" = "#DC3220"
    )
  ) +
  # set point colors to match bar fill colors
  scale_color_manual(
    values = c(
      "Native" = "#005AB5",
      "Invasive" = "#DC3220"
    ),
    guide = "none"
  ) +
  # add plot title, axis labels, and legend label
  labs(
    title = "Change in Mean % Cover of Key Species Post-burn",
    x = "Species",
    y = "Change in Mean % Cover\n(percentage points)",
    fill = NULL
  ) +
  # non default theme
  theme_bw() +
  # adjust title, axis, and legend appearance
  theme(
    plot.title = element_text(
      size = 18,
      face = "bold",
      hjust = 0.5
    ),
    axis.title = element_text(size = 14),
    axis.text.x = element_text(
      size = 13,
      face = "italic"
    ),
    axis.text.y = element_text(size = 12),
    legend.position = c(0.15, 0.82),
    legend.background = element_rect(fill = "transparent")
  )

# display plot
percent_point_cover_change_plot
```

```{r}
#| label: saving percent change plot
#| message: false
#| echo: false

# saving ggplot
ggsave(
  # file path and name
  filename = here("outputs", "figure_4.png"),
  plot = percent_point_cover_change_plot,
  width = 8,
  height = 5,
  dpi = 300
)
```


## Plot 4: Mean percent cover of native grasses after disturbance
```{r}

```

## Exploratory Visualization 1: Change in mean percent cover of key species after sheep grazing
```{r}
#| label: calculating sheep grazing change in mean percent cover
#| message: false
#| echo: false

## Change in percentage points of focal species percent cover
## 2024 = pre-grazing
## 2025 = post-grazing

# creating a table for the four focal species used in this figure
focal_species_sheep <- tibble(
  psoc = c(
    "Stipa_pulchra",
    "Medicago_polymorpha",
    "Melilotus_indicus",
    "Festuca_perennis"
  ),
  # shortened species labels for plotting
  species_label = c(
    "S. pulchra",
    "M. polymorpha",
    "M. indicus",
    "F. perennis"
  ),
  # classify each species as native or invasive
  status = c(
    "Native",
    "Invasive",
    "Invasive",
    "Invasive"
  )
)

# create a new object from grassland explicit 0s
focal_cover_sheep <- gl_explicit0 |>
  # keep only 2024 and 2025 observations for the four focal species
  filter(
    year %in% c(2024, 2025),
    psoc %in% focal_species_sheep$psoc
  ) |>
  # keep only the columns needed for calculating change
  select(year, transect_name, psoc, mean_pc) |>
  # add species labels and native or invasive status
  left_join(focal_species_sheep, by = "psoc")

# create a new object from focal cover sheep
focal_change_transect_sheep_pp <- focal_cover_sheep |>
  # convert year values into separate columns for comparison
  pivot_wider(
    names_from = year,
    values_from = mean_pc,
    names_prefix = "year_"
  ) |>
  # subtract pre grazing cover from post grazing cover
  mutate(
    percent_point_change = year_2025 - year_2024
  )

# create a new object from the new calculated percent point change object
focal_change_summary_sheep_pp <- focal_change_transect_sheep_pp |>
  # group by species and status
  group_by(psoc, species_label, status) |>
  # calculate the mean percent point change and sample size
  summarize(
    mean_percent_point_change = mean(percent_point_change, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) |>
  # set plotting order for species and status categories
  mutate(
    species_label = factor(
      species_label,
      levels = c(
        "S. pulchra",
        "M. polymorpha",
        "M. indicus",
        "F. perennis"
      )
    ),
    status = factor(status, levels = c("Native", "Invasive"))
  )
```

```{r}
#| label: plotting sheep grazing change in mean percent cover
#| message: false
#| echo: false
#| fig-width: 8
#| fig-height: 4
#| fig-cap: "Change in mean percent cover of four key grassland species on the NCOS Mesa from 2024, pre-grazing, to 2025, post-grazing. Bars show the average percent-point change across transects, individual points show transect-level changes, and the gray horizontal line represents no change; blue indicates the native species Stipa pulchra, while red indicates invasive species. As an exploratory visualization, this figure suggests that S. pulchra changed only slightly after grazing, while invasive species showed mixed responses, with M. polymorpha and F. perennis increasing on average and M. indicus decreasing slightly."

# create a ggplot with baselayer as the calculated percent point changes
percent_point_cover_change_sheep_plot <- ggplot(
  focal_change_summary_sheep_pp,
  # x axis = focal species, y axis = mean percent point change,
  # fill = native or invasive status
  aes(x = species_label, y = mean_percent_point_change, fill = status)
) +
  # bars show mean percent point change from 2024 to 2025
  geom_col(width = 0.6) +
  # jittered points show transect level percent point changes
  geom_jitter(
    data = focal_change_transect_sheep_pp,
    aes(x = species_label, y = percent_point_change, color = status),
    width = 0.08,
    height = 0,
    size = 2,
    alpha = 0.6,
    inherit.aes = FALSE
  ) +
  # horizontal reference line at 0 showing no change between years
  geom_hline(
    yintercept = 0,
    color = "gray50"
  ) +
  # add rounded mean percent point change labels above or below columns
  geom_text(
    aes(
      label = round(mean_percent_point_change, 0),
      y = if_else(
        mean_percent_point_change >= 0,
        mean_percent_point_change + 2,
        mean_percent_point_change - 2
      )
    ),
    size = 5,
    color = "black"
  ) +
  # set fill colors for native and invasive bar categories
  scale_fill_manual(
    values = c(
      "Native" = "#005AB5",
      "Invasive" = "#DC3220"
    )
  ) +
  # set point colors to match bar fill colors
  scale_color_manual(
    values = c(
      "Native" = "#005AB5",
      "Invasive" = "#DC3220"
    ),
    guide = "none"
  ) +
  # add plot title, axis labels, and legend label
  labs(
    title = "Change in Mean % Cover of Key Species Post-grazing",
    x = "Species",
    y = "Change in Mean % Cover\n(percentage points)",
    fill = NULL
  ) +
  # non default theme
  theme_bw() +
  # adjust title, axis, and legend appearance
  theme(
    plot.title = element_text(
      size = 18,
      face = "bold",
      hjust = 0.5
    ),
    axis.title = element_text(size = 14),
    axis.text.x = element_text(
      size = 13,
      face = "italic"
    ),
    axis.text.y = element_text(size = 12),
    legend.position = c(0.15, 0.82),
    legend.background = element_rect(fill = "transparent")
  )

# display plot
percent_point_cover_change_sheep_plot
```


## Exploratory Visualization 2: Effect of annual precipitation on native and invasive percent cover

```{r}
#| label: data
#| message: false
#| echo: false

weather <- read_csv(here("data", "raw", "NOAA-weather-data.csv"))
# creating new object from weather
weather_clean <- weather |> 
  # cleaning column names
  clean_names() |> 
  # making sure date is in POSIXct format
  mutate(date = mdy(date)) |> 
  # selecting columns of interest
  select(date, prcp, tmax, tmin) |> 
  # extracting month and year from date column
  mutate(month = month(date),
         year = year(date)) |> 
  # assigning water year
  mutate(wy = case_when(
    # when month is October or later, assign value of year + 1
    month >= 10 ~ year + 1,
    # when month is September or earlier, assign value of existing year
    .default = year))
```

```{r}
#| label: data
#| message: false
#| echo: false


# new obj
 tmap <- weather_clean |>
  # group by year
  group_by(year) |>
  # calculate total annual precip
  summarize(total_precip = sum(prcp, na.rm = TRUE)) |> 
  # ungroup 
  ungroup()
```

```{r}
#| label: data
#| message: false
#| echo: false


# new obj
precip_veg <- full_join(
  # left
  x = tmap, 
  # right
  y = grasses_summary,
  # shared column
  by = c("year" = "year") 
) |> 
  # selecting relevant columns
  select(year, grand_mean_pc, cover_category, total_precip, se) |> 
  # remove NAs and year 2020
  filter(!is.na(grand_mean_pc), !is.na(total_precip), year != 2020)
```

```{r}
#| label: precip vs grand mean % cover
#| fig-width: 8
#| fig-height: 4
#| fig-cap: " Changes in grand mean percent cover for native and non-native cover categories on NCOS Mesa from 2021-2025. Individual points show grand mean percent cover for each year. Dark grey line shows trendline; blue line represents native cover; orange represents non-native cover; dotted line represents total annual precipitation. As an exploratory visualization, this figure suggests that precipitation is a factor that increases non-native percent cover while precipitation doesn't increase native percent cover."


TAPnat <- ggplot(data = precip_veg,
       # x axis is year, y axis is grand mean % cover, color is cover category
       mapping = aes(x = year,
                     y = grand_mean_pc,
                     color = cover_category)) +
  # first layer: points for year and grand mean % cover
  geom_point() +
  # second layer: lines to connect points
  geom_line() +
  # trend line
  geom_smooth(method = "lm", se = TRUE, aes(linetype = "Trend Line", color = "Trend Line")) +
  # add precipitation as secondary y axis scaled to grand mean % cover range
  geom_line(aes(y = total_precip / max(precip_veg$total_precip, na.rm = TRUE) *
                    max(precip_veg$grand_mean_pc, na.rm = TRUE),
                color = "Total Precipitation"),
            linetype = "dashed") +
  # facet wrap to display cover category
  facet_wrap(~cover_category) +
  # add color for cover categories, precipitation, and trend line
  scale_color_manual(name = "Legend",
                     values = c("NATIVE COVER" = "#005AB5",
                                "NON-NATIVE COVER" = "#DC3220",
                                "Total Precipitation" = "#228B22",
                                "Trend Line" = "gray30"),
                     labels = c("Native", "Non-native", "Total Precipitation", "Trend Line")) +
  scale_linetype_manual(name = "Legend",
                        values = c("Trend Line" = "solid")) +
  # merge into one legend, suppress separate linetype legend
  guides(color = guide_legend(override.aes = list(linetype = c("solid", "solid", "dashed", "solid"),
                                                   shape = c(16, 16, NA, NA))),
         linetype = "none") +
  # dual y axes
  scale_y_continuous(
    name = "Grand Mean Percent Cover (%)",
    sec.axis = sec_axis(
      ~ . * max(precip_veg$total_precip, na.rm = TRUE) /
            max(precip_veg$grand_mean_pc, na.rm = TRUE),
      name = "Total Annual Precipitation (in)"
    )
  ) +
  # labels
  labs(
    title = "Precipitation and grand mean % cover of native and nonnative cover",
    x = "Year") +
  theme_bw()

TAPnat

```


# Plan for elective
- The general concept is we are making an ArcGIS StoryMap presenting our NCOS Mesa vegetation analysis. The StoryMap walks through site context, the 2023 prescribed burn, methods, results, and an embedded interactive dashboard where users can explore transect photo points and a spatial percent cover heatmap across the Mesa Perennial Grassland.

- The full structure is built out and mostly populated. The completed sections include site context, burn background, methods, and results with all four figures and the t-test table embedded. Remaining work includes building the ArcGIS Dashboard with photo point popups and the interpolated heatmap layer, embedding it into the StoryMap, and finalizing the discussion and references sections.

- [Link to StoryMap](https://arcg.is/1rfKbf5)

