---
title: "Timeline Check in"
author: "Lukas Lescano, Garrett Mellinger, Minh Tri Ngo"
date: "2026-05-22"
format: pdf
message: false
warning: false
---

```{r}
#| label: packages
#| message: false
#| echo: false

library(tidyverse)
library(janitor)
library(ggrepel)
library(cowplot)
library(here)
library(ggtext)
```


```{r}
#| label: data
#| message: false
#| echo: false

# read in data
veg <- read_csv(here("data", "raw", "veg.csv"))

metadata <- read_csv(here("data", "raw", "vp_veg_metadata.csv"))
```


```{r}
#| label: adding target species list
#| message: false
#| echo: false

poaceae_species <- c(
  "Bromus_hordeaceus",
  "Festuca_perennis",
  "Stipa_pulchra",
  "Polypogon_monspeliensis",
  "Brachypodium_distachyon",
  "Festuca_myuros",
  "Avena_barbata",
  "Hordeum_marinum",
  "Avena_fatua",
  "Parapholis_incurva",
  "Bromus_diandrus",
  "Distichlis_spicata",
  "Bromus_catharticus",
  "Hordeum_murinum",
  "Hordeum_brachyantherum_brachy.",
  "Polypogon_viridis",
  "Cortaderia_selloana",
  "Festuca_bromoides",
  "Cynodon_dactylon",
  "Poa_annua",
  "Festuca_microstachys",
  "Crypsis_schoenoides",
  "Polypogon_interruptus",
  "Hordeum_brachyantherum",
  "Bromus_madritensis",
  "Elymus_triticoides"
)

native_forbs <- c(
  "Plantago_erecta",
  "Sisyrinchium_bellum",
  "Sidalcea_malviflora",
  "Eryngium_vaseyi",
  "Acmispon_maritimus",
  "Acmispon_wrangelianus",
  "Madia_sativa",
  "Psilocarphus_tennulus",
  "Psilocarphus_brevissimus",
  "Spergularia_macrotheca",
  "Centromadia_parryi_australis",
  "Lepidium_nitidum",
  "Stebbinsoseris_heterocarpa",
  "Laennecia_coulteri"
)

exclude_other <- c(
  "Bare Ground",
  "Natural Thatch",
  "unlisted",
  "moss",
  "boulder_rock",
  "biocrust",
  "Baccharis_pilularis",
  "Juncus_bufonius",
  "Juncus_phaeocephalus",
  "Eleocharis_macrostachya",
  "Eleocharis_acicularis",
  "Eleocharis_palustris"
)
```


```{r}
#| label: creating species lookup
#| message: false
#| echo: false

species_lookup <- veg |>
  # keep only GL transects at ncos
  filter(str_detect(transect_name, "GL"), site == "ncos") |>
  # remove unlisted since it maps to multiple categories
  filter(psoc != "unlisted") |>
  # return one unique row per species-category combination
  distinct(psoc, cover_category) |>
  # sort alphabetically by species for readability
  arrange(psoc)
```


```{r}
#| label: creating explicit0 dataframe
#| message: false
#| echo: false

# create a new object from veg
gl_explicit0 <- veg |> 
  # filter all occurrences where transect_name contains VP
  filter(str_detect(transect_name, "GL"),
         site == "ncos") |> 
  # not including cover_category because we will join it back after complete()
  select(-cover_category) |> 
  # complete all possible combinations of year, pool, psoc and distance
  complete(year, transect_name, psoc, transect_distance,
           # fill values of percent cover with 0
           fill = list(percent_cover = 0)) |>
  # group by year, pool, and psoc
  group_by(year, transect_name, psoc) |>
  # calculate total percent cover across all quads
  summarize(sum_pc = sum(percent_cover, na.rm = TRUE)) |>
  # ungroup the data frame
  ungroup() |>
  # create a new column called year_pool from year and transect name
  unite("year_pool", year, transect_name, remove = TRUE) |> 
  # join with metadata data frame
  left_join(metadata, by = "year_pool") |> 
  # calculate mean percent cover from total percent cover/number of quadrats
  mutate(mean_pc = sum_pc/num_quad) |> 
  # re-attach cover_category from lookup using psoc as key
  left_join(species_lookup, by = "psoc") |> 
  # excluding "GL-11" because it was unburned
  filter(transect_name != "GL-11")
```


```{r}
#| label: creating grasses dataframe
#| message: false
#| echo: false

# creating grasses data frame from gl_explicit0
grasses <- gl_explicit0 |> 
  # filtering to include poaceae grasses species
  filter(psoc %in% poaceae_species,
         # filtering to include only native and non-native cover categories
         cover_category %in% c("NATIVE COVER", "NON-NATIVE COVER"))
```

```{r}
#| label: creating grasses_summary for time series plot
#| message: false
#| echo: false

grasses_summary <- grasses |> 
  # grouping by year and cover category
  group_by(year, cover_category) |> 
  # calculating average per transect mean_pc across all transects and species within each group
  summarise(grand_mean_pc = mean(mean_pc, na.rm = TRUE),
            se = sd(mean_pc, na.rm = TRUE) / sqrt(n())) |> 
  # remove grouping
  ungroup()
```


```{r}
#| label: creating grasses transect data frame
#| message: false
#| echo: false

# creating new object from grasses
grasses_transect <- grasses |>
  # grouping by year, transect, and conver category
  group_by(year, transect_name, cover_category) |>
  # calculating mean pc
  summarize(mean_pc = mean(mean_pc, na.rm = TRUE)) |>
  # removing groups
  ungroup()
```

# Ideas for analysis

## Questions

- How does the percent cover of native grasses vary over time compared to the percent cover of exotic vegetation on the NCOS Mesa Perennial Grasslands?

    - How does intentional disturbance influence this relationship?
    - How do key individual species, including *Stipa pulchra*, *Medicago polymorpha*, *Melilotus indicus*, and *Festuca perennis*, vary in percent cover before and after intentional disturbance?

## Statistical analysis
- statistical analysis to address at least one question with at least one analysis coded and run


# Background

We assess whether the 2023 cultural burn is meeting its restoration goals of promoting native grassland species. This study may help to inform future restoration management decisions regarding intentional disturbance at NCOS and could have implications which extend to the management or ecological understanding of other coastal grassland systems. These questions are relevant because Santa Barbara coastal grasslands are fire adapted (Saglimbeni, 2024, pp. 1-2), but are now heavily affected by invasive species. Studying the NCOS burn helps show whether cultural burning can support native species like Stipa pulchra while reducing nonnative grassland invaders.


# Visualizations

## Plot 1: Native and exotic percent cover over time
```{r}

```

## Plot 2: Sweet and bur clover abundance over time
```{r}

```

## Plot 3: Change in mean percent cover of key species after disturbance
```{r}
#| label: calculating change in mean percent cover
#| message: false
#| warning: false
#| echo: false

## Change in percentage points of focal species percent cover
## 2023 = pre-burn
## 2024 = post-burn

# creating a table for the four focal species used in this figure
focal_species <- tibble(
  psoc = c(
    "Stipa_pulchra",
    "Medicago_polymorpha",
    "Melilotus_indicus",
    "Festuca_perennis"
  ),
  # shortened species labels for plotting
  species_label = c(
    "S. pulchra",
    "M. polymorpha",
    "M. indicus",
    "F. perennis"
  ),
  # classify each species as native or invasive
  status = c(
    "Native",
    "Invasive",
    "Invasive",
    "Invasive"
  )
)

# create a new object from grassland explicit 0s
focal_cover <- gl_explicit0 |>
  # keep only 2023 and 2024 observations for the four focal species
  filter(
    year %in% c(2023, 2024),
    psoc %in% focal_species$psoc
  ) |>
  # keep only the columns needed for calculating change
  select(year, transect_name, psoc, mean_pc) |>
  # add species labels and native or invasive status
  left_join(focal_species, by = "psoc")

# create a new object from focal cover
focal_change_transect_pp <- focal_cover |>
  # convert year values into separate columns for comparison
  pivot_wider(
    names_from = year,
    values_from = mean_pc,
    names_prefix = "year_"
  ) |>
  # subtract pre-burn cover from post-burn cover
  mutate(
    percent_point_change = year_2024 - year_2023
  )

# create a new object from the new calculated percent point change object
focal_change_summary_pp <- focal_change_transect_pp |>
  # group by species and status
  group_by(psoc, species_label, status) |>
  # calculate the mean percent point change and sample size
  summarize(
    mean_percent_point_change = mean(percent_point_change, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) |>
  # set plotting order for species and status categories
  mutate(
    species_label = factor(
      species_label,
      levels = c(
        "S. pulchra",
        "M. polymorpha",
        "M. indicus",
        "F. perennis"
      )
    ),
    status = factor(status, levels = c("Native", "Invasive"))
  )
```

```{r}
#| label: plotting change in mean percent cover
#| message: false
#| echo: false
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "Change in mean percent cover of four key grassland species on the NCOS Mesa from 2023, pre-burn, to 2024, one year post-burn. Bars represent the average percent-point change in cover across transects for each species, while individual points show transect-level percent-point changes; the gray horizontal line represents no change, with values above zero indicating increased cover and values below zero indicating decreased cover. Blue indicates the native species, Stipa pulchra, and red indicates invasive species: Medicago polymorpha, Melilotus indicus, and Festuca perennis. Overall, *S. pulchra* and *F. perennis* decreased after the burn, *M. indicus* increased, and *M. polymorpha* showed little average change. This figure connects to our main question by showing that intentional disturbance did not affect all native and invasive species in the same way."

# create a ggplot with baselayer as the calculated percent point changes
percent_point_cover_change_plot <- ggplot(
  focal_change_summary_pp,
  # x axis = focal species, y axis = mean percent point change,
  # fill = native or invasive status
  aes(x = species_label, y = mean_percent_point_change, fill = status)
) +
  # bars show mean percent point change from 2023 to 2024
  geom_col(width = 0.6) +
  # jittered points show transect level percent point changes
  geom_jitter(
    data = focal_change_transect_pp,
    aes(x = species_label, y = percent_point_change, color = status),
    width = 0.08,
    height = 0,
    size = 2,
    alpha = 0.6,
    inherit.aes = FALSE
  ) +
  # horizontal reference line at 0 showing no change between years
  geom_hline(
    yintercept = 0,
    color = "gray50"
  ) +
  # add rounded mean percent point change labels above or below columns
  geom_text(
    aes(
      label = round(mean_percent_point_change, 0),
      y = if_else(
        mean_percent_point_change >= 0,
        mean_percent_point_change + 2.9,
        mean_percent_point_change - 2.4
      )
    ),
    size = 5,
    color = "black"
  ) +
  # set fill colors for native and invasive bar categories
  scale_fill_manual(
    values = c(
      "Native" = "#005AB5",
      "Invasive" = "#DC3220"
    )
  ) +
  # set point colors to match bar fill colors
  scale_color_manual(
    values = c(
      "Native" = "#005AB5",
      "Invasive" = "#DC3220"
    ),
    guide = "none"
  ) +
  # add plot title, axis labels, and legend label
  labs(
    title = "Change in Mean % Cover of Key Species Post-burn",
    x = "Species",
    y = "Change in Mean % Cover\n(percentage points)",
    fill = NULL
  ) +
  # non default theme
  theme_bw() +
  # adjust title, axis, and legend appearance
  theme(
    plot.title = element_text(
      size = 18,
      face = "bold",
      hjust = 0.5
    ),
    axis.title = element_text(size = 14),
    axis.text.x = element_text(
      size = 13,
      face = "italic"
    ),
    axis.text.y = element_text(size = 12),
    legend.position = c(0.15, 0.82),
    legend.background = element_rect(fill = "transparent")
  )

# display plot
percent_point_cover_change_plot
```

## Plot 4: Mean percent cover of native grasses after disturbance
```{r}

```

## Exploratory Visualization 1: Change in mean percent cover of key species after sheep grazing
```{r}
#| label: calculating sheep grazing change in mean percent cover
#| message: false
#| echo: false

## Change in percentage points of focal species percent cover
## 2024 = pre-grazing
## 2025 = post-grazing

# creating a table for the four focal species used in this figure
focal_species_sheep <- tibble(
  psoc = c(
    "Stipa_pulchra",
    "Medicago_polymorpha",
    "Melilotus_indicus",
    "Festuca_perennis"
  ),
  # shortened species labels for plotting
  species_label = c(
    "S. pulchra",
    "M. polymorpha",
    "M. indicus",
    "F. perennis"
  ),
  # classify each species as native or invasive
  status = c(
    "Native",
    "Invasive",
    "Invasive",
    "Invasive"
  )
)

# create a new object from grassland explicit 0s
focal_cover_sheep <- gl_explicit0 |>
  # keep only 2024 and 2025 observations for the four focal species
  filter(
    year %in% c(2024, 2025),
    psoc %in% focal_species_sheep$psoc
  ) |>
  # keep only the columns needed for calculating change
  select(year, transect_name, psoc, mean_pc) |>
  # add species labels and native or invasive status
  left_join(focal_species_sheep, by = "psoc")

# create a new object from focal cover sheep
focal_change_transect_sheep_pp <- focal_cover_sheep |>
  # convert year values into separate columns for comparison
  pivot_wider(
    names_from = year,
    values_from = mean_pc,
    names_prefix = "year_"
  ) |>
  # subtract pre grazing cover from post grazing cover
  mutate(
    percent_point_change = year_2025 - year_2024
  )

# create a new object from the new calculated percent point change object
focal_change_summary_sheep_pp <- focal_change_transect_sheep_pp |>
  # group by species and status
  group_by(psoc, species_label, status) |>
  # calculate the mean percent point change and sample size
  summarize(
    mean_percent_point_change = mean(percent_point_change, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) |>
  # set plotting order for species and status categories
  mutate(
    species_label = factor(
      species_label,
      levels = c(
        "S. pulchra",
        "M. polymorpha",
        "M. indicus",
        "F. perennis"
      )
    ),
    status = factor(status, levels = c("Native", "Invasive"))
  )
```

```{r}
#| label: plotting sheep grazing change in mean percent cover
#| message: false
#| echo: false
#| fig-width: 8
#| fig-height: 4
#| fig-cap: "Change in mean percent cover of four key grassland species on the NCOS Mesa from 2024, pre-grazing, to 2025, post-grazing. Bars show the average percent-point change across transects, individual points show transect-level changes, and the gray horizontal line represents no change; blue indicates the native species Stipa pulchra, while red indicates invasive species. As an exploratory visualization, this figure suggests that S. pulchra changed only slightly after grazing, while invasive species showed mixed responses, with M. polymorpha and F. perennis increasing on average and M. indicus decreasing slightly."

# create a ggplot with baselayer as the calculated percent point changes
percent_point_cover_change_sheep_plot <- ggplot(
  focal_change_summary_sheep_pp,
  # x axis = focal species, y axis = mean percent point change,
  # fill = native or invasive status
  aes(x = species_label, y = mean_percent_point_change, fill = status)
) +
  # bars show mean percent point change from 2024 to 2025
  geom_col(width = 0.6) +
  # jittered points show transect level percent point changes
  geom_jitter(
    data = focal_change_transect_sheep_pp,
    aes(x = species_label, y = percent_point_change, color = status),
    width = 0.08,
    height = 0,
    size = 2,
    alpha = 0.6,
    inherit.aes = FALSE
  ) +
  # horizontal reference line at 0 showing no change between years
  geom_hline(
    yintercept = 0,
    color = "gray50"
  ) +
  # add rounded mean percent point change labels above or below columns
  geom_text(
    aes(
      label = round(mean_percent_point_change, 0),
      y = if_else(
        mean_percent_point_change >= 0,
        mean_percent_point_change + 2,
        mean_percent_point_change - 2
      )
    ),
    size = 5,
    color = "black"
  ) +
  # set fill colors for native and invasive bar categories
  scale_fill_manual(
    values = c(
      "Native" = "#005AB5",
      "Invasive" = "#DC3220"
    )
  ) +
  # set point colors to match bar fill colors
  scale_color_manual(
    values = c(
      "Native" = "#005AB5",
      "Invasive" = "#DC3220"
    ),
    guide = "none"
  ) +
  # add plot title, axis labels, and legend label
  labs(
    title = "Change in Mean % Cover of Key Species Post-grazing",
    x = "Species",
    y = "Change in Mean % Cover\n(percentage points)",
    fill = NULL
  ) +
  # non default theme
  theme_bw() +
  # adjust title, axis, and legend appearance
  theme(
    plot.title = element_text(
      size = 18,
      face = "bold",
      hjust = 0.5
    ),
    axis.title = element_text(size = 14),
    axis.text.x = element_text(
      size = 13,
      face = "italic"
    ),
    axis.text.y = element_text(size = 12),
    legend.position = c(0.15, 0.82),
    legend.background = element_rect(fill = "transparent")
  )

# display plot
percent_point_cover_change_sheep_plot
```

## Exploratory Visualization 2: Effect of annual precipitation on native and invasive percent cover
```{r}

```


# Plan for elective
- the general concept

- a description of the progress you have made

- a link, photo, or some visual of your elective

