---
title: "Visualizations"
author: "Lukas Lescano, Garrett Mellinger, Minh Tri Ngo"
date: "2026-05-18"
format: pdf
---

```{r}
#| label: packages
#| message: false
#| echo: false

library(tidyverse)
library(janitor)
library(ggrepel)
library(cowplot)
library(here)
library(ggtext)
```

```{r}
#| label: data
#| message: false
#| echo: false

# read in data
veg <- read_csv(here("data", "raw", "veg.csv"))

metadata <- read_csv(here("data", "raw", "vp_veg_metadata.csv"))

```

```{r}
#| label: adding target species list
#| message: false
#| echo: false

poaceae_species <- c(
  "Bromus_hordeaceus",
  "Festuca_perennis",
  "Stipa_pulchra",
  "Polypogon_monspeliensis",
  "Brachypodium_distachyon",
  "Festuca_myuros",
  "Avena_barbata",
  "Hordeum_marinum",
  "Avena_fatua",
  "Parapholis_incurva",
  "Bromus_diandrus",
  "Distichlis_spicata",
  "Bromus_catharticus",
  "Hordeum_murinum",
  "Hordeum_brachyantherum_brachy.",
  "Polypogon_viridis",
  "Cortaderia_selloana",
  "Festuca_bromoides",
  "Cynodon_dactylon",
  "Poa_annua",
  "Festuca_microstachys",
  "Crypsis_schoenoides",
  "Polypogon_interruptus",
  "Hordeum_brachyantherum",
  "Bromus_madritensis",
  "Elymus_triticoides"
)

native_forbs <- c(
  "Plantago_erecta",
  "Sisyrinchium_bellum",
  "Sidalcea_malviflora",
  "Eryngium_vaseyi",
  "Acmispon_maritimus",
  "Acmispon_wrangelianus",
  "Madia_sativa",
  "Psilocarphus_tennulus",
  "Psilocarphus_brevissimus",
  "Spergularia_macrotheca",
  "Centromadia_parryi_australis",
  "Lepidium_nitidum",
  "Stebbinsoseris_heterocarpa",
  "Laennecia_coulteri"
)

exclude_other <- c(
  "Bare Ground",
  "Natural Thatch",
  "unlisted",
  "moss",
  "boulder_rock",
  "biocrust",
  "Baccharis_pilularis",
  "Juncus_bufonius",
  "Juncus_phaeocephalus",
  "Eleocharis_macrostachya",
  "Eleocharis_acicularis",
  "Eleocharis_palustris"
)
```

```{r}
#| label: creating species lookup
#| message: false
#| echo: false

species_lookup <- veg |>
  # keep only GL transects at ncos
  filter(str_detect(transect_name, "GL"), site == "ncos") |>
  # remove unlisted since it maps to multiple categories
  filter(psoc != "unlisted") |>
  # return one unique row per species-category combination
  distinct(psoc, cover_category) |>
  # sort alphabetically by species for readability
  arrange(psoc)
```

```{r}
#| label: creating pools_explicit0 dataframe
#| message: false
#| echo: false

# create a new object from veg
gl_explicit0 <- veg |> 
  # filter all occurrences where transect_name contains VP
  filter(str_detect(transect_name, "GL"),
         site == "ncos") |> 
  # not including cover_category because we will join it back after complete()
  select(-cover_category) |> 
  # complete all possible combinations of year, pool, psoc and distance
  complete(year, transect_name, psoc, transect_distance,
           # fill values of percent cover with 0
           fill = list(percent_cover = 0)) |>
  # group by year, pool, and psoc
  group_by(year, transect_name, psoc) |>
  # calculate total percent cover across all quads
  summarize(sum_pc = sum(percent_cover, na.rm = TRUE)) |>
  # ungroup the data frame
  ungroup() |>
  # create a new column called year_pool from year and transect name
  unite("year_pool", year, transect_name, remove = TRUE) |> 
  # join with metadata data frame
  left_join(metadata, by = "year_pool") |> 
  # calculate mean percent cover from total percent cover/number of quadrats
  mutate(mean_pc = sum_pc/num_quad) |> 
  # re-attach cover_category from lookup using psoc as key
  left_join(species_lookup, by = "psoc") |> 
  # excluding "GL-11" because it was unburned
  filter(transect_name != "GL-11")
```

```{r}
#| label: creating grasses dataframe
#| message: false
#| echo: false

# creating grasses data frame from gl_explicit0
grasses <- gl_explicit0 |> 
  # filtering to include poaceae grasses species
  filter(psoc %in% poaceae_species,
         # filtering to include only native and non-native cover categories
         cover_category %in% c("NATIVE COVER", "NON-NATIVE COVER"))
```

```{r}
#| label: creating grasses_summary for time series plot
#| message: false
#| echo: false

grasses_summary <- grasses |> 
  # grouping by year and cover category
  group_by(year, cover_category) |> 
  # calculating average per transect mean_pc across all transects and species within each group
  summarise(grand_mean_pc = mean(mean_pc, na.rm = TRUE),
            se = sd(mean_pc, na.rm = TRUE) / sqrt(n())) |> 
  # remove grouping
  ungroup()
```

```{r}
#| label: creating grasses transect data frame
#| message: false
#| echo: false

# creating new object from grasses
grasses_transect <- grasses |>
  # grouping by year, transect, and conver category
  group_by(year, transect_name, cover_category) |>
  # calculating mean pc
  summarize(mean_pc = mean(mean_pc, na.rm = TRUE)) |>
  # removing groups
  ungroup()
```

```{r}
#| label: plot-aesthetics
#| message: false
#| echo: false

colors <- c(
  "#0077BB",
  "#33BBEE",
  "#009988",
  "#EE7733",
  "#CC3311",
  "#EE3377",
  "#AA4499"
)

theme_set(
  theme_bw() +
    theme(text = element_text(size = 14))
)
```

## Plot 1: Timeseries

```{r}
#| label: timeseries-plot
#| message: false
#| echo: false
#| fig-width: 10
#| fig-height: 5
#| fig-cap: "Mean percent cover of native and non-native grasses along the Mesa (GL) transects at the North Campus Open Space (NCOS) from 2021 to 2025. Filled circles represent the grand mean across transects and open circles represent individual transect means. Shaded ribbons show ±1 standard error. Dashed vertical lines indicate the prescribed burn (Fall 2023) and rotational sheep grazing (Fall 2024) treatment events."

TS_plot <- ggplot() +
  # individual transect observations as open circles
  geom_jitter(data = grasses_transect,
             aes(x = year, y = mean_pc, color = cover_category),
             shape = 16, alpha = 0.4, size = 2, width = 0.1, height = 0.1) +
  # ribbon showing ±1 SE around the grand mean; low alpha to reduce visual noise
  geom_ribbon(data = grasses_summary,
              aes(x = year, ymin = grand_mean_pc - se, ymax = grand_mean_pc + se,
                  fill = cover_category), alpha = 0.1) +
  # line connecting the grand means across years
  geom_line(data = grasses_summary,
            aes(x = year, y = grand_mean_pc, color = cover_category)) +
  # grand mean as filled circle
  geom_point(data = grasses_summary,
             aes(x = year, y = grand_mean_pc, color = cover_category),
             shape = 18, size = 5) +
  # dashed line marking the prescribed burn intentional disturbance event
  geom_vline(xintercept = 2023.5, linetype = "dashed", color = "black") +
  # dashed line marking the rotational grazing intentional disturbance event
  geom_vline(xintercept = 2024.5, linetype = "dashed", color = "black") +
  # text label for burn event; hjust pushes label right of the line
  annotate("text", x = 2023.5, y = 10,
           label = "Prescribed Burn", hjust = -0.1, size = 3.5) +
  # text label for grazing event; hjust pushes label right of the line
  annotate("text", x = 2024.5, y = 10,
           label = "Sheep Grazing", hjust = -0.1, size = 3.5) +
  # setting color colors; colors must match to merge legends
  scale_color_manual(name = "Cover Type",
                   values = c("NATIVE COVER" = "#005AB5",
                              "NON-NATIVE COVER" = "#DC3220"),
                   labels = c("Native", "Non-native")) +
  # setting fill colors; colors must match to merge legends
  scale_fill_manual(name = "Cover Type",
                  values = c("NATIVE COVER" = "#005AB5",
                             "NON-NATIVE COVER" = "#DC3220"),
                  labels = c("Native", "Non-native")) +
  # labeling the axes
  labs(x = "Year",
       y = "Mean Percent Cover",
       color = "Cover Type",
       # consider dropping title since figure caption conveys the same information
       title = "Native vs. Non-Native Grass Cover Over Time") +
  # moving legend position to the top-left of figure panel
  theme(legend.position = c(0.15, 0.82),
        # transparent legend background so underlying data points are visible
        legend.background = element_rect(fill = "transparent"))

# displaying plot
TS_plot
```


```{r}
#| label: plot 2 wrangling
#| message: false
#| echo: false

# creating new object from gl_explicit0
clover <- gl_explicit0 |>
  # filter to clover species of interest
  filter(psoc %in% c("Melilotus_indicus", "Medicago_polymorpha"))
```

```{r}
#| label: creating clover summary data frame
#| message: false
#| echo: false

# creating new object from clover
clover_summary <- clover |>
  # grouping by year and psoc
  group_by(year, psoc) |>
  # calculating grand mean percent cover
  summarize(grand_mean_pc = mean(mean_pc, na.rm = TRUE),
            # calculating standard error
            se = sd(mean_pc, na.rm = TRUE) / sqrt(n())) |>
  # removing groups
  ungroup()
```

## Plot 2: Sweet and Bur Clover Over Time

```{r}
#| label: creating clover plot
#| message: false
#| echo: false
#| fig-width: 10
#| fig-height: 5
#| fig-cap: "Mean percent cover of bur clover (*M. polymorpha*) and sweet clover (*M. indicus*) along the Mesa (GL) transects at the North Campus Open Space (NCOS) from 2021 to 2025. Filled circles represent the grand mean across transects and open circles represent individual transect means. Shaded ribbons show ±1 standard error. Dashed vertical lines indicate the prescribed burn (Fall 2023) and rotational sheep grazing (Fall 2024) treatment events. Field observations suggested *M. polymorpha* increased following the prescribed burn and *M. indicus* following sheep grazing, a pattern reflected in the monitoring data."

# creating new plot
clover_plot <- ggplot() +
  # individual transect observations as open circles
  geom_jitter(data = clover,
              aes(x = year, y = mean_pc, color = psoc),
              shape = 16, alpha = 0.4, size = 2,
              width = 0.1, height = 0.1) +
  # ribbon showing standard error around the grand mean
  geom_ribbon(data = clover_summary,
              aes(x = year, ymin = grand_mean_pc - se, ymax = grand_mean_pc + se,
                  fill = psoc), alpha = 0.1) +
  # line connecting grand means across years
  geom_line(data = clover_summary,
            aes(x = year, y = grand_mean_pc, color = psoc)) +
  # grand mean as filled circle
  geom_point(data = clover_summary,
             aes(x = year, y = grand_mean_pc, color = psoc),
             shape = 18, size = 5) +
  # dashed line marking prescribed burn between 2023 and 2024 surveys
  geom_vline(xintercept = 2023.5, linetype = "dashed", color = "black") +
  # dashed line marking rotational sheep grazing between 2024 and 2025 surveys
  geom_vline(xintercept = 2024.5, linetype = "dashed", color = "black") +
  # text label for burn event
  annotate("text", x = 2023.5, y = 42,
           label = "Prescribed Burn", hjust = -0.1, size = 3.5) +
  # text label for grazing event
  annotate("text", x = 2024.5, y = 42,
           label = "Sheep Grazing", hjust = -0.1, size = 3.5) +
  # setting color colors; colors must match to merge legends
  scale_color_manual(name = "Species",
                   values = c("Melilotus_indicus" = "#E66100",
                              "Medicago_polymorpha" = "#5D3A9B"),
                   labels = c("<i>M. indicus</i> (sweet clover)", 
                              "<i>M. polymorpha</i> (bur clover)")) +
  # setting fill colors; colors must match to merge legends
  scale_fill_manual(name = "Species",
                  values = c("Melilotus_indicus" = "#E66100",
                             "Medicago_polymorpha" = "#5D3A9B"),
                  labels = c("<i>M. indicus</i> (sweet clover)", 
                             "<i>M. polymorpha</i> (bur clover)")) +
  # labeling axes
  labs(x = "Year",
       y = "Mean Percent Cover",
       title = "Sweet and Bur Clover Cover Over Time") +
  # enable HTML markdown to render in legend text to display the italic species names
  theme(legend.text = element_markdown(),
        # moving legend position to the top-left of figure panel
        legend.position = c(0.15, 0.82),
        # transparent legend background so underlying data points are visible
        legend.background = element_rect(fill = "transparent"))

# displaying plot
clover_plot
```

## Plot 3: Change in Mean Percent Cover of Key Species Diverging Bar Chart

```{r}
#| label: calculating change in mean percent cover
#| message: false
#| warning: false
#| echo: false

## Change in percentage points of focal species percent cover
## 2023 = pre-burn
## 2024 = post-burn

# creating a table for the four focal species used in this figure
focal_species <- tibble(
  psoc = c(
    "Stipa_pulchra",
    "Medicago_polymorpha",
    "Melilotus_indicus",
    "Festuca_perennis"
  ),
  # shortened species labels for plotting
  species_label = c(
    "S. pulchra",
    "M. polymorpha",
    "M. indicus",
    "F. perennis"
  ),
  # classify each species as native or invasive
  status = c(
    "Native",
    "Invasive",
    "Invasive",
    "Invasive"
  )
)

# create a new object from grassland explicit 0s
focal_cover <- gl_explicit0 |>
  # keep only 2023 and 2024 observations for the four focal species
  filter(
    year %in% c(2023, 2024),
    psoc %in% focal_species$psoc
  ) |>
  # keep only the columns needed for calculating change
  select(year, transect_name, psoc, mean_pc) |>
  # add species labels and native or invasive status
  left_join(focal_species, by = "psoc")

# create a new object from focal cover
focal_change_transect_pp <- focal_cover |>
  # convert year values into separate columns for comparison
  pivot_wider(
    names_from = year,
    values_from = mean_pc,
    names_prefix = "year_"
  ) |>
  # subtract pre-burn cover from post-burn cover
  mutate(
    percent_point_change = year_2024 - year_2023
  )

# create a new object from the new calculated percent point change object
focal_change_summary_pp <- focal_change_transect_pp |>
  # group by species and status
  group_by(psoc, species_label, status) |>
  # calculate the mean percent point change and sample size
  summarize(
    mean_percent_point_change = mean(percent_point_change, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) |>
  # set plotting order for species and status categories
  mutate(
    species_label = factor(
      species_label,
      levels = c(
        "S. pulchra",
        "M. polymorpha",
        "M. indicus",
        "F. perennis"
      )
    ),
    status = factor(status, levels = c("Native", "Invasive"))
  )
```


```{r}
#| label: plotting change in mean percent cover
#| message: false
#| echo: false
#| fig-width: 8
#| fig-height: 4
#| fig-cap: "Change in mean percent cover of four key grassland species on the NCOS Mesa from 2023, pre-burn, to 2024, one year post-burn. Bars represent the average percent-point change in cover across transects for each species, while individual points show transect-level percent-point changes; the gray horizontal line represents no change, with values above zero indicating increased cover and values below zero indicating decreased cover. Blue indicates the native species, Stipa pulchra, and red indicates invasive species: Medicago polymorpha, Melilotus indicus, and Festuca perennis. Overall, *S. pulchra* and *F. perennis* decreased after the burn, *M. indicus* increased, and *M. polymorpha* showed little average change. This figure connects to our main question by showing that intentional disturbance did not affect all native and invasive species in the same way."

# create a ggplot with baselayer as the calculated percent point changes
percent_point_cover_change_plot <- ggplot(
  focal_change_summary_pp,
  # x axis = focal species, y axis = mean percent point change,
  # fill = native or invasive status
  aes(x = species_label, y = mean_percent_point_change, fill = status)
) +
  # bars show mean percent point change from 2023 to 2024
  geom_col(width = 0.6) +
  # jittered points show transect level percent point changes
  geom_jitter(
    data = focal_change_transect_pp,
    aes(x = species_label, y = percent_point_change, color = status),
    width = 0.08,
    height = 0,
    size = 2,
    alpha = 0.6,
    inherit.aes = FALSE
  ) +
  # horizontal reference line at 0 showing no change between years
  geom_hline(
    yintercept = 0,
    color = "gray50"
  ) +
  # add rounded mean percent point change labels above or below columns
  geom_text(
    aes(
      label = round(mean_percent_point_change, 0),
      y = if_else(
        mean_percent_point_change >= 0,
        mean_percent_point_change + 2.9,
        mean_percent_point_change - 2.4
      )
    ),
    size = 5,
    color = "black"
  ) +
  # set fill colors for native and invasive bar categories
  scale_fill_manual(
    values = c(
      "Native" = "#005AB5",
      "Invasive" = "#DC3220"
    )
  ) +
  # set point colors to match bar fill colors
  scale_color_manual(
    values = c(
      "Native" = "#005AB5",
      "Invasive" = "#DC3220"
    ),
    guide = "none"
  ) +
  # add plot title, axis labels, and legend label
  labs(
    title = "Change in Mean % Cover of Key Species Post-burn",
    x = "Species",
    y = "Change in Mean % Cover\n(percentage points)",
    fill = NULL
  ) +
  # non default theme
  theme_bw() +
  # adjust title, axis, and legend appearance
  theme(
    plot.title = element_text(
      size = 18,
      face = "bold",
      hjust = 0.5
    ),
    axis.title = element_text(size = 14),
    axis.text.x = element_text(
      size = 13,
      face = "italic"
    ),
    axis.text.y = element_text(size = 12),
    legend.position = c(0.15, 0.82),
    legend.background = element_rect(fill = "transparent")
  )

# display plot
percent_point_cover_change_plot
```


## Plot 4: Disturbance boxplot

```{r}
#| label: creating dataframe for before and after treatment
#| echo: false
#| message: false


# create new object
nfs <- grasses_transect |> 
  filter(year %in% c("2023","2024","2025"),
         cover_category == "NATIVE COVER") |> 
  group_by(transect_name, year) |> 
  mutate(before_after = factor(
    case_when( 
      year == 2023 ~ "No Treatment",
      year == 2024 ~ "Fire",
      year == 2025 ~ "Sheep"
    ),
    levels = c("No Treatment", "Fire", "Sheep")
  ))

```



```{r}
#| label: boxplot creation
#| echo: false
#| message: false
#| fig-width: 10
#| fig-height: 5

#creating new plot
treatments_bp <- ggplot(data = nfs,
                  # x-axis: Treatment Type
                  # y-axis: % Cover Native Grasses
                  # fill geometries by (before and after)
                  mapping = aes(x = before_after,
                                y = mean_pc,
                                fill = before_after)) +
  # layer 3: jitter 
  geom_jitter(width = 0.2, alpha=1.0, color = "#5D3A9B")+
  # boxplot (and narrowing width)
  geom_boxplot(width = 0.5) +
  # filling geometries
  scale_fill_manual(values = c("No Treatment" = "#E1Be6A",
                               "Fire" = "#E66100",
                               "Sheep" = "#1A85FF")) +
  # relabelling x- and y-axes, adding title
  labs(x = "Treatment Type",
       y = "% Cover Native Grasses",
       title = "Mean % Cover by Treatment Type 2021-2025") +
  # theme
  theme_bw()+
  # remove legend from theme
  theme(legend.position = "none")

treatments_bp

```

## Exploratory Plot 1

## Exploratory Plot 2

## Citations

### 1. Evaluating Sheep Grazing as a Strategy for Restoring the North Campus Open Space (NCOS) Grasslands (eScholarship Cheadle Center Poster)

[Link to paper #1](https://escholarship.org/uc/item/9jp6r7zs)

Notes:

- Confirmed focal species selection; our four key species align with those identified in the poster
- Inspired figure 3 percent change bar chart from their visualization approach
- Adopted paired t-test framework comparing pre and post treatment mean percent cover
- Study is site-specific to NCOS making it directly applicable to our analysis

### 2. North Campus Open Space Restoration Project Monitoring Report: Year 7, December 2024

[Link to paper #2](https://escholarship.org/uc/item/16p046mb)

Notes:

- Used for prescribed burn timing and treatment details
- Provided site context for the Mesa (GL) perennial grassland transects
- Primary data source for vegetation monitoring records used in analysis
- Covers 2017-2024 monitoring period providing pre-treatment baseline context for our 2021-2025 analysis window
- Confirms NCOS restoration scope including upland native grassland habitats relevant to our mesa transects

### 3. Trajectories of vegetation-based indicators used to assess wetland restoration progress

[Link to paper #3](https://esajournals.onlinelibrary.wiley.com/doi/full/10.1890/08-1371.1)

Notes:

- Non-monotonic restoration trajectories -- supports native cover declining post-burn not meaning restoration is failing
- Justifies need for longer-term monitoring to assess native grass recovery
- Note wetland vs. grassland limitation when citing

### 4. A decade-long study of repeated prescription burning in California native grassland restoration

[Link to paper #4](https://onlinelibrary.wiley.com/doi/full/10.1111/rec.13939)

Notes:

- Directly supports prescribed burn as a restoration tool in Stipa pulchra dominated California grasslands; same species and system as our study
- Contextualize the post-burn decline in our native grass cover; the paper shows S. pulchra recovery takes multiple years, consistent with the drop we see in 2024
- Non-native annual grasses reduced post-burn in their study, which partially aligns with our non-native grass decline post-burn
- Support longer-term monitoring; recovery patterns only became clear over 5+ years, reinforcing that our 2021-2025 window may not capture full recovery
- Note that their study used repeated burns on the same plots while ours is a single burn event; acknowledge this difference when citing

### 5. Evaluating Prescribed Fire Effect on Medusa Head and Other Invasive Plants in Coastal Prairie at Point Pinole

[Link to paper #5](https://cdn.wildapricot.com/76181/resources/Documents/Grasslands%20Journal/Grasslands%20Articles/Research%20articles/Bartolome%20et%20al%20Fire%20Effect%20on%20Medusa%20Head_Winter%202019.pdf?version=1629934104000&Policy=eyJTdGF0ZW1lbnQiOiBbeyJSZXNvdXJjZSI6Imh0dHBzOi8vY2RuLndpbGRhcHJpY290LmNvbS83NjE4MS9yZXNvdXJjZXMvRG9jdW1lbnRzL0dyYXNzbGFuZHMlMjBKb3VybmFsL0dyYXNzbGFuZHMlMjBBcnRpY2xlcy9SZXNlYXJjaCUyMGFydGljbGVzL0JhcnRvbG9tZSUyMGV0JTIwYWwlMjBGaXJlJTIwRWZmZWN0JTIwb24lMjBNZWR1c2ElMjBIZWFkX1dpbnRlciUyMDIwMTkucGRmP3ZlcnNpb249MTYyOTkzNDEwNDAwMCIsIkNvbmRpdGlvbiI6eyJEYXRlTGVzc1RoYW4iOnsiQVdTOkVwb2NoVGltZSI6MTc3OTQ3NzMxMH0sIklwQWRkcmVzcyI6eyJBV1M6U291cmNlSXAiOiIwLjAuMC4wLzAifX19XX0_&Signature=CFXo6jtgwF-Hs0EuVv26HjcxmFU-HcNid4ACqszuUeDT-kuA2eBh0NJlq3bEnGRSnVwqoKh5hc6xEWh7NlcFCkB-6ZLeWZsPD018BWjfvf-6kQIlLyWrRawJal8tmYKVq2p2ECKhNmrfzenS62HtEkn2-DunErLCJeHSwiD6Y-EWzMDniR2RAAmgo0c8S-ixY3FZfPdmqhZwk2YAwcHYyoI-UQ6gQ5BUtkQwlApaPGOEcGUavBoa5SllpVXrj0dbItmjoTWDrTqUPUEnqRlrwycKitNxzIfDB56Ga5dMELhfjY6vzZ6eNGWzY57sXHxCfSpKdE0K8LsPARub3U2b6g__&Key-Pair-Id=K27MGQSHTHAGGF)

Notes:

- Coastal prairie California grassland setting makes this more ecologically comparable to NCOS mesa than most burn studies
- Stipa pulchra and Brachypodium distachyon also appear in our study providing direct species-level comparability
- Supports prescribed burn as a management tool for invasive grasses in California coastal grasslands specifically
- Note timing difference; their burn was conducted in June while ours was Fall 2023, which could affect results since burn timing influences which species are most impacted
- Methodology is comparable; transect-based vegetation monitoring pre and post burn

### 6. Prescribed Burn Impacts on California Coastal Prairie Seed Banks

[Link to paper #6](https://bioone.org/journals/madro%c3%b1o/volume-73/issue-1/0024-9637-260004/PRESCRIBED-BURN-IMPACTS-ON-CALIFORNIA-COASTAL-PRAIRIE-SEED-BANKS/10.3120/0024-9637-260004.full)

Notes:

- Central Coast California coastal prairie setting is directly comparable to NCOS mesa; strongest geographic match of any paper so far
- Supports the finding that single prescribed burns are insufficient to reduce non-native grass abundance; contextualizes why we still see high non-native cover post-burn in our data
- Fall 2024 sampling timeline overlaps directly with our post-burn monitoring period
- Contextualizes the mixed burn effects we see; non-native grasses not declining as expected is consistent with their findings
- The point that non-native plants colonize disturbed landscapes faster than natives supports the post-burn increase we see in clover species
- Note that their study focuses on seed banks while ours measures above-ground percent cover; complementary approaches to the same question
- Useful for introducing prescribed burn management context in our background/introduction section

### 7. Grazing and Fire Management for Native Perennial Grass Restoration in California Grasslands

[Link to paper #7](https://ucanr.edu/sites/default/files/2010-09/51815.pdf)

Notes:

- Directly addresses combined grazing and prescribed fire as restoration tools for native perennial grass restoration in California grasslands; most directly relevant to our previous dual treatment design
- Cite to support the use of sheep grazing to reduce invasive annual grass seed banks and thatch, directly applicable to NCOS grazing treatment rationale
- Stipa pulchra is a focal species in this paper; supports our use of it as a native grass indicator
- Notes that perennial grasses show reduced seed production the first year post-fire but recover by year two or three; contextualizes our 2024 post-burn S. pulchra decline as an expected short-term response
- Recommends burning every third or fourth year since non-native annual grass composition returns to pre-fire status in about three years; useful for discussing management implications
- Seeding of annual clover noted as potentially compatible with established perennials but competitive effects on native seedlings unknown; directly relevant to our clover increase post-treatment
- Cite in methods/discussion to justify the combined burn and grazing treatment approach used at NCOS